---
title: "AUC Tables for Multiple Predictors: aucmat Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AUC Tables for Multiple Predictors: aucmat Vignette}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(aucmat)
```

# Introduction

**aucmat** is an R package for computing ROC AUC tables for multiple predictors using the `pROC` package. It provides robust missing-data handling with several imputation strategies and optional ranking of predictors by AUC.

This vignette demonstrates the main features of aucmat, using example data.

# Example Data

Let's create a simple synthetic dataset:

```{r}
set.seed(123)
n <- 200
df <- data.frame(
  Outcome = sample(0:1, n, replace = TRUE),
  Predictor1 = rnorm(n, mean = 50, sd = 10),
  Predictor2 = rnorm(n, mean = 30, sd = 5),
  Predictor3 = rnorm(n, mean = 70, sd = 20)
)

# Introduce missing values
df$Predictor1[sample(n, 20)] <- NA
df$Predictor2[sample(n, 15)] <- NA
df$Predictor3[sample(n, 10)] <- NA

head(df)
```

# Basic Usage

To compute AUC for each predictor:

```{r}
result <- aucmat(
  data = df,
  outcome_col = "Outcome",
  predictors = c("Predictor1", "Predictor2", "Predictor3"),
  na_strategy = "none",  # Remove rows with missing values
  rank = TRUE            # Rank predictors by AUC
)
print(result)
```

# Using Different Missing Data Strategies

aucmat supports multiple ways to handle missing data. For example, you can impute missing values using the median stratified by outcome:

```{r}
result_median_by_class <- aucmat(
  data = df,
  outcome_col = "Outcome",
  predictors = c("Predictor1", "Predictor2", "Predictor3"),
  na_strategy = "median_by_class"
)
print(result_median_by_class)
```

Other available strategies include: `mean`, `median`, `constant`, `halfmin`, `quantile`, `knn`, `mice`, and `missForest`.

# Confidence Intervals

aucmat computes confidence intervals for AUC using either the DeLong method or bootstrap:

```{r}
result_boot <- aucmat(
  data = df,
  outcome_col = "Outcome",
  predictors = c("Predictor1", "Predictor2", "Predictor3"),
  ci_method = "bootstrap"
)
print(result_boot)
```

# Tidy Output and Ranking

aucmat outputs a tidy data frame. You can optionally rank predictors by their AUC values:

```{r}
result_ranked <- aucmat(
  data = df,
  outcome_col = "Outcome",
  predictors = c("Predictor1", "Predictor2", "Predictor3"),
  rank = TRUE
)
print(result_ranked)
```

# Advanced: Imputation with knn, mice, or missForest

If you have the required packages installed, you can use advanced imputation strategies:

```{r eval=FALSE}
# Requires mice and missForest packages to be installed
result_mice <- aucmat(df, "Outcome", c("Predictor1", "Predictor2", "Predictor3"), na_strategy = "mice")
result_missForest <- aucmat(df, "Outcome", c("Predictor1", "Predictor2", "Predictor3"), na_strategy = "missForest")
```

# Summary

aucmat provides a simple and flexible workflow for evaluating multiple predictors using ROC AUC, with strong support for missing data and ranking options. See the package documentation for further details and options.
