% ===========================================================================
%  aucmat: Matrix-First ROC Analysis
%  Lucas VH Huynh-Tran
%  July 2026
% ===========================================================================

\documentclass[11pt,a4paper]{article}

% ---- Packages ----------------------------------------------------------
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

% ---- Hyperref setup ----------------------------------------------------
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

% ---- Code environment --------------------------------------------------
\DefineVerbatimEnvironment{CodeBlock}{Verbatim}{
  fontsize=\small,
  frame=single,
  numbers=left,
  numbersep=5pt,
  xleftmargin=2em
}

% ---- Title / Authors ---------------------------------------------------
\title{\textbf{aucmat: Matrix-First ROC Analysis\\[4pt]
        \large Scalable ROC Analysis for High-Dimensional Biomarker Matrices}}

\author{Lucas VH Huynh-Tran%
  \thanks{Correspondence: \href{https://github.com/vanhungtran}{github.com/vanhungtran}}%
  \\[4pt]
  \small R package version 0.2.0%
}

\date{July 24, 2026}

% ---- Document ----------------------------------------------------------
\begin{document}

\maketitle

\begin{abstract}
\noindent
Receiver operating characteristic (ROC) analysis is the dominant framework
for evaluating continuous biomarkers against binary outcomes in diagnostic
medicine, prognostic modeling, and molecular screening.  Despite decades of
methodological development, no widely-used R package integrates matrix-wide
screening, direction-preserving AUC, multiplicity adjustment, paired
comparisons, stability analysis, cross-validation, and panel construction
into a single reproducible workflow.  We present \texttt{aucmat}, an R
package whose matrix-first design accepts a numeric matrix $\mathbf{X} \in
\mathbb{R}^{n \times p}$ and a binary outcome vector $\mathbf{y} \in
\{0,1\}^n$ and returns a complete screening table---point estimates, standard
errors, confidence intervals, raw and multiplicity-adjusted $p$-values, and
diagnostic metadata---for every biomarker in a single call.  \texttt{aucmat}
provides direction-preserving AUC reporting, DeLong and bootstrap inference,
Benjamini--Hochberg/Holm/Bonferroni multiplicity adjustment, paired
biomarker comparisons with non-inferiority and equivalence (TOST) tests, an
omnibus Wald test for three or more correlated AUCs, bootstrap rank-stability
analysis, a two-stage Hurdle-AUC model for zero-inflated data, multivariable
panel construction with honestly nested cross-validation, a multivariate
simulation engine with calibration diagnostics, and power/sample-size
computation.  Screens of 100, 1{,}000, and 10{,}000 biomarkers ($n=500$,
DeLong inference) complete in 0.54, 5.02, and 59.76 seconds.  DeLong 95\%
CI coverage across 27 simulation conditions ranges 0.898--0.957, converging
to the nominal 0.95 with increasing sample size.  We demonstrate the package
on plasma Olink proteomics from the BeatMG trial of rituximab for myasthenia
gravis, where \texttt{aucmat} recovers the expected pharmacodynamic signature
of anti-CD20 B-cell depletion---six of eight on-treatment-significant
proteins are canonical B-cell surface markers---while a null baseline screen
rules out technical artifacts.

\smallskip
\noindent\textbf{Keywords:} ROC analysis, AUC, biomarker screening, DeLong
test, multiplicity adjustment, R package, proteomics, Hurdle model
\end{abstract}

% ==========================================================================
\section{Introduction}
% ==========================================================================

Receiver operating characteristic (ROC) analysis is the dominant framework
for evaluating continuous biomarkers against binary outcomes in diagnostic
medicine, prognostic modeling, and molecular screening
\citep{Hanley1982,DeLong1988,Pepe2004}.  The area under the ROC curve (AUC)
summarizes the probability that a randomly chosen positive subject has a
higher biomarker value than a randomly chosen negative subject.  Its
interpretability---0.5 for chance, 1.0 for perfect separation---has made it
the default metric in omics-scale biomarker discovery \citep{Robin2011}.

Modern molecular profiling technologies routinely measure thousands of
features on a few hundred subjects.  A proteomics panel may contain 5{,}000
proteins; a metabolomics screen 2{,}000 metabolites; a transcriptomics
experiment 20{,}000 genes.  The core statistical task---``which biomarkers
discriminate cases from controls?''---requires computing and testing an AUC
for each feature, adjusting for multiplicity, comparing top candidates, and
assessing the stability of the rankings under sampling variability.  Despite
decades of methodological development, no widely-used R package integrates
these steps into a single, reproducible workflow.

The existing R ecosystem provides excellent point solutions.  \textbf{pROC}
\citep{Robin2011} is the gold standard for single-biomarker ROC inference:
it computes DeLong confidence intervals using the Sun \& Xu (2014)
$\mathcal{O}(N \log N)$ algorithm \citep{Sun2014}, tests paired or unpaired
AUC differences via \texttt{roc.test()}, handles partial AUC with bootstrap
inference, provides binormal and density-smoothed ROC curves, computes
covariance between paired AUCs via \texttt{cov()}, and supports multiclass
AUC via Hand \& Till \citep{Hand2001}.  However, pROC operates one biomarker
at a time---screening a matrix of 10{,}000 features requires an explicit
per-column loop.  \textbf{ROCR} \citep{Sing2005} offers flexible
visualization with 28 performance measures and cross-validation averaging
but provides no statistical inference (no confidence intervals, no hypothesis
tests for AUC differences).  \textbf{precrec} \citep{Saito2017} adds
precision-recall curves and fast C++-based AUC computation, with confidence
bands for multiple test sets and functions for partial AUC
(\texttt{part()}, \texttt{pauc()}) and AUC confidence intervals
(\texttt{auc\_ci()}), but provides no DeLong inference, no multiplicity
adjustment, and no paired comparison tests.  \textbf{colAUC} from
\texttt{caTools} \citep{Tuszynski2024} computes column-wise AUCs efficiently
via the Wilcoxon rank-sum test and supports multiclass AUC via column
means---but provides no inference whatsoever (no standard errors, confidence
intervals, or $p$-values).  \textbf{dtComb} \citep{Yerlitas2025} provides
142 methods for combining two diagnostic markers into a composite score,
with 30+ cutoff optimization methods, resampling, and \texttt{caret}
integration---but is limited to exactly two biomarkers and does not perform
matrix-wide screening.  \textbf{cancerclass} \citep{Budczies2024} builds
validated classifiers from high-dimensional molecular data using
nearest-centroid prediction with multiple random validation, but does not
offer DeLong-based screening or bootstrap stability analysis.

\texttt{aucmat} fills this gap.  Its design philosophy is
\textit{matrix-first}: the primary function \texttt{aucmat(X, y)} accepts a
numeric matrix $\mathbf{X} \in \mathbb{R}^{n \times p}$ and a binary outcome
vector $\mathbf{y} \in \{0,1\}^n$, and returns a complete screening table
with point estimates, standard errors, confidence intervals, raw and
multiplicity-adjusted $p$-values, and diagnostic metadata for every biomarker
in a single call.  Downstream functions for comparison, stability,
cross-validation, panel construction, and power analysis consume the same S3
object, preserving outcome encoding and inferential settings throughout.

% ==========================================================================
\section{Methods}
% ==========================================================================

\subsection{Direction-preserving AUC}

For biomarker $X$ and binary outcome $Y \in \{0,1\}$, the AUC equals the
Mann--Whitney $U$ probability:
\begin{equation}\label{eq:auc_mw}
\text{AUC} = P(X_i > X_j \mid Y_i = 1, Y_j = 0)
\end{equation}
\citep{Hanley1982}.  Most ROC software reports $\max(\text{AUC}, 1 -
\text{AUC})$, silently reversing the biomarker to force the reported value
above 0.5.  \texttt{aucmat} deliberately preserves the observed direction,
reporting three values per biomarker:
$\text{AUC}_{\text{raw}}$ (the Mann--Whitney estimate on the original numeric
scale, which may be below 0.5),
$\text{AUC}_{\text{strength}} = 0.5 + |\text{AUC}_{\text{raw}} - 0.5|$
(for ranking), and \texttt{effect\_direction} (\texttt{higher\_in\_positive}
or \texttt{lower\_in\_positive}).  This distinction matters biologically: a
tumor suppressor gene with $\text{AUC}_{\text{raw}} = 0.15$ discriminates
just as strongly as an oncogene with $\text{AUC}_{\text{raw}} = 0.85$, but
with the opposite biological interpretation.

\subsection{DeLong variance and joint covariance}

The DeLong variance estimator \citep{DeLong1988,Sun2014} uses placement
values
\begin{equation}\label{eq:V10}
V_{10}(X_i) = \frac{1}{n_0} \sum_{j:Y_j=0}
\bigl[\mathbf{1}(X_j < X_i) + \tfrac{1}{2} \cdot \mathbf{1}(X_j = X_i)\bigr]
\end{equation}
for positive subjects and
\begin{equation}\label{eq:V01}
V_{01}(X_j) = \frac{1}{n_1} \sum_{i:Y_i=1}
\bigl[\mathbf{1}(X_i > X_j) + \tfrac{1}{2} \cdot \mathbf{1}(X_i = X_j)\bigr]
\end{equation}
for negative subjects.  Let
$S_{10} = (n_1 - 1)^{-1} \sum_{i \in \mathcal{I}_1}
(V_{10}(X_i) - \bar{V}_{10})^2$ and
$S_{01} = (n_0 - 1)^{-1} \sum_{j \in \mathcal{I}_0}
(V_{01}(X_j) - \bar{V}_{01})^2$ be the sample variances of the placement
values.  The DeLong variance estimator is
\begin{equation}\label{eq:delong_var}
\widehat{\text{Var}}(\widehat{\text{AUC}}) = \frac{S_{10}}{n_1} +
\frac{S_{01}}{n_0}.
\end{equation}

For $p$ biomarkers on common subjects, the $p \times p$ joint covariance
matrix
\begin{equation}\label{eq:joint_cov}
\widehat{\bm{\Sigma}} = n_1^{-1}\,\widehat{\text{Cov}}(\mathbf{V}_{10})
+ n_0^{-1}\,\widehat{\text{Cov}}(\mathbf{V}_{01})
\end{equation}
supports an omnibus Wald test of
$H_0: \text{AUC}_1 = \cdots = \text{AUC}_p$.
Let $\mathbf{C}$ be a $(p-1) \times p$ contrast matrix (default: comparing
biomarkers $2,\ldots,p$ against biomarker 1).  The Wald statistic
\begin{equation}\label{eq:wald}
W = (\mathbf{C}\hat{\bm{\theta}})^\top
(\mathbf{C}\widehat{\bm{\Sigma}}\mathbf{C}^\top)^{-1}
(\mathbf{C}\hat{\bm{\theta}}) \sim \chi^2_{r}
\end{equation}
uses the effective rank $r$ of the contrast covariance matrix, determined
via eigenvalue tolerance.  A generalized inverse is used on the stable
subspace only; rank-zero or numerically unstable cases return
\texttt{status = "non\_testable\_singular"}.  For $p = 2$, $W$ equals the
squared paired-DeLong $z$-statistic within numerical tolerance, serving as
an internal consistency check.

\subsection{Multiplicity adjustment and stability}

Screening $p$ biomarkers produces $p$ hypothesis tests.
\texttt{aucmat} applies Benjamini--Hochberg \citep{Benjamini1995}, Holm, or
Bonferroni correction via \texttt{stats::p.adjust()}, reporting adjusted
$p$-values as $q$-values.  For equivalence (TOST) \citep{Schuirmann1987},
the reported $p_{\text{equiv}} = \max(p_{\text{lower}}, p_{\text{upper}})$
and multiplicity adjustment applies to this combined value.

Bootstrap rank-stability analysis (\texttt{auc\_stability()}) resamples the
data $B$ times with replacement (stratified by outcome), recomputes the full
screening for each replicate, and reports median/IQR rank distributions,
top-$k$ selection probabilities, and pairwise co-selection
frequencies---quantifying how reproducible the ranking is under sampling
variability \citep{Efron1994}.

\subsection{Hurdle-AUC for zero-inflated data}

Standard AUC assumes continuous biomarkers following a shifted normal
distribution per class.  This assumption breaks on zero-inflated data
(scRNA-seq, microbiome, detection-limited assays) where 60\%+ of values are
exactly zero.  Under the standard model, zeros would need to arise from the
tail of a normal distribution below zero---biologically implausible.  The
Hurdle-AUC uses a two-stage model:

\textbf{Stage 1---Zero hurdle:}
$P(X_j = 0 \mid Y) = \operatorname{logistic}(\beta_0 + \beta_1 Y)$ models
the binary expression decision via logistic regression.

\textbf{Stage 2---Magnitude:} Standard AUC on the non-zero values captures
magnitude discrimination among expressed observations.

The composite score $S_i$ for observation $i$ is:
$S_i = \hat{p}_i$ if $X_{ij} = 0$, and
$S_i = 0.5 \cdot \hat{p}_i + 0.5 \cdot \tilde{x}_{ij}$ if $X_{ij} > 0$,
where $\hat{p}_i$ is the predicted probability of being non-zero from
Stage~1, and $\tilde{x}_{ij}$ is the normalized biomarker value.

Simulation confirms recovery from standard AUC $\approx 0.53$ to Hurdle-AUC
$\approx 0.90$ on zero-inflated biomarkers by acknowledging that ``expression
on/off'' and ``expression level when on'' are separate biological processes.

\subsection{Multivariable panel construction}

Screening ranks biomarkers individually; combining several into one score
is a distinct problem, addressed by \texttt{fit\_auc\_panel()}.  Four of its
six methods are ridge, lasso, elastic-net, and unpenalized logistic
regression on centred and scaled predictors, the penalized variants fit via
\texttt{glmnet} \citep{Friedman2010}.  A fifth, \texttt{su\_liu}, is the
closed-form linear combination that maximizes AUC under the common-covariance
multivariate normal model \citep{Su1993}: for class-conditional means
$\bm{\mu}_1, \bm{\mu}_0$ and pooled within-class covariance $\bm{\Sigma}$,
the AUC-maximizing direction is
\begin{equation}\label{eq:su_liu}
\mathbf{w} = \bm{\Sigma}^{-1} (\bm{\mu}_1 - \bm{\mu}_0)
\end{equation}
---the same direction as Fisher's linear discriminant, and a natural
counterpart to the binormal model already used by \texttt{aucmat}'s
simulation engine.  Because $\bm{\Sigma}$ is frequently ill-conditioned when
biomarkers are collinear or $p$ approaches $n$, the pooled covariance is
shrunk toward its diagonal,
\begin{equation}\label{eq:shrinkage}
\bm{\Sigma}_\lambda = (1-\lambda)\bm{\Sigma} +
\lambda\,\operatorname{diag}(\bm{\Sigma})
\end{equation}
for a user-set $\lambda \in [0,1]$, with a reciprocal-condition-number guard
that fails informatively rather than returning an unstable solution.  A
sixth, \texttt{unweighted}, sums direction-aligned, standardized biomarkers
with equal weight and fits no parameters at all---an assumption-light
baseline for judging whether the fitted methods earn their added complexity.

All six methods share one \textbf{nested} cross-validation procedure for the
reported out-of-fold AUC: centering, scaling, and (for the penalized methods)
penalty selection are re-derived from each outer fold's training portion
alone, including for the final full-data model's own internal penalty search,
so nothing derived from a held-out fold ever contributes to the model
evaluated on it.  \texttt{compare\_panel\_auc()} reuses the DeLong or
bootstrap paired-comparison engine from \texttt{compare\_auc()} to test the
panel's honest cross-validated score against each individual biomarker used
to build it, on the same subjects, and \texttt{plot\_roc\_panel()} plots the
corresponding ROC curves together.

\subsection{Multivariate simulation}

Under the equal-variance binormal model, $X_k \mid Y \sim
\mathcal{N}(\mu_{kY}, \sigma^2)$ with
$\delta_k = (\mu_{1k} - \mu_{0k})/\sigma = \sqrt{2} \cdot
\Phi^{-1}(\text{AUC}_k)$.  For $p$ biomarkers with correlation matrix
$\mathbf{R}$, the class-conditional distribution is $\mathbf{X} \mid Y \sim
\mathcal{N}_p(\bm{\mu}_Y, \mathbf{R})$, where $\bm{\mu}_0$ and $\bm{\mu}_1$
are centred so the $n$-weighted grand mean is zero.
\texttt{simulate\_auc\_matrix()} generates data under this model with four
parametrized correlation structures (user, exchangeable, AR(1), block) and
feasibility diagnostics with optional nearest positive-definite projection.

\texttt{simulate\_auc\_copula()} uses a two-phase approach: Phase~1 applies a
Gaussian copula (rank-transform to uniform, probit to normal scores) that
preserves the target correlation structure; Phase~2 applies iterative mean
perturbation with decaying step size to fine-tune each biomarker's AUC.  This
separates correlation control (copula) from discriminative signal
(perturbation), avoiding the AUC--correlation tradeoff inherent in sequential
binormal methods.

\texttt{simulate\_hurdle\_auc()} implements the two-stage hurdle model:
Bernoulli zeros with class-specific rates, and log-normal magnitudes for
expressed values with binormal-calibrated mean separation, clipped to
$[0, \infty)$.

\texttt{validate\_simulation()} repeats a specification across independent
draws and reports bias, RMSE, Monte Carlo standard error, and
target-interval hit rates for both AUCs and pairwise correlations---a single
draw is never evidence that a simulator is calibrated.

% ==========================================================================
\section{Implementation}
% ==========================================================================

\texttt{aucmat} is implemented in pure R ($\sim$6{,}900 lines across 37
source files) with minimized dependencies.  The primary function
\texttt{aucmat()} performs six steps in sequence:
\begin{enumerate}
  \item Input validation (matrix dimensions, column names, outcome levels,
        infinite values);
  \item Outcome normalization via \texttt{.normalize\_binary\_y()} (accepting
        logical, numeric 0/1, factor, or character outcomes);
  \item Matrix-wide AUC computation via the Mann--Whitney rank formulation;
  \item DeLong or bootstrap inference via \texttt{.apply\_inference()} with
        configurable alternative hypothesis (\texttt{two.sided},
        \texttt{greater}, \texttt{less});
  \item Multiplicity adjustment via \texttt{.adjust\_pvalues()};
  \item Result assembly with ranking by discrimination strength and
        feature-level warnings for small class counts or high missingness.
\end{enumerate}

All primary functions return S3 objects with \texttt{print()},
\texttt{summary()}, \texttt{plot()}, and (where appropriate)
\texttt{as.data.frame()} and \texttt{subset()} methods.  The package stores
outcome encoding, positive-class designation, and sample masks in the
returned object, eliminating the need for downstream functions to re-infer
factor direction or re-derive class counts.

Structured error classes (inheriting from \texttt{error}) allow programmatic
handling of common failures: \texttt{aucmat\_invalid\_outcome},
\texttt{aucmat\_invalid\_correlation}, \texttt{aucmat\_infeasible\_targets},
\texttt{aucmat\_insufficient\_sample}, and
\texttt{aucmat\_resampling\_failure}.  An RNG-safety mechanism
(\texttt{with\_seed()}) saves the global \texttt{.Random.seed}, executes the
computation, and restores the previous state on exit, including when
\texttt{.Random.seed} was previously absent and when an error occurs.

The package imports \texttt{pROC} (ROC curve objects and smoothed ROC),
\texttt{mvtnorm} (multivariate normal generation), \texttt{ggplot2}
(visualization), and \texttt{rlang} (tidy evaluation).  Optional
\texttt{Suggests} include \texttt{GGally} (pairs plots for simulated data),
\texttt{glmnet} (penalized panel models), \texttt{matrixStats} (fast
column-wise ranks), \texttt{knitr} and \texttt{rmarkdown} (vignette
building), and \texttt{ggrepel} (non-overlapping plot labels).  The complete
source, documentation, and vignettes are available at
\url{https://vanhungtran.github.io/aucmat}.

% ==========================================================================
\section{Usage}
% ==========================================================================

\subsection{Basic screening}

\begin{CodeBlock}
library(aucmat)

sim <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.3, structure = "exchangeable"
)
X <- as.matrix(sim$data[, 1:4])
y <- sim$data$truth

fit <- aucmat(X, y, ci = "delong", adjust = "BH")
print(fit)
summary(fit)
plot_auc_rank(fit, n_label = 4)
\end{CodeBlock}

\subsection{Paired comparisons with hypothesis testing}

\begin{CodeBlock}
# Two-sided superiority (default: H0: delta = 0)
compare_auc(fit, X, y, reference = "X1")

# Non-inferiority: H0: AUC_a <= AUC_b - margin
compare_auc(fit, X, y, reference = "X1",
  hypothesis = "noninferiority", margin = 0.05)

# Equivalence (TOST): H0: |AUC_a - AUC_b| >= margin
compare_auc(fit, X, y, reference = "X1",
  hypothesis = "equivalence", margin = 0.15, adjust = "BH")

# Omnibus Wald test: H0: AUC_1 = AUC_2 = ... = AUC_p
compare_auc_global(fit, X, y, biomarkers = paste0("X", 1:4))

# Single-biomarker DeLong test against chance
roc_test(X[, 1], y)
\end{CodeBlock}

\subsection{Hurdle-AUC for zero-inflated data}

\begin{CodeBlock}
# scRNA-seq: 55% zeros in controls, 25% in cases
sim <- simulate_hurdle_auc(
  n = 400, prevalence = 0.3,
  target_hurdle_aucs = c(0.85, 0.72),
  zero_rate_neg = c(0.55, 0.80),
  zero_rate_pos = c(0.25, 0.70)
)
fit_hur <- hurdle_auc(as.matrix(sim$data[, 1:2]), sim$data$truth)
print(fit_hur)
# Standard AUC ~0.53 recovers to Hurdle AUC ~0.90
\end{CodeBlock}

\subsection{Cross-validation, panel construction, power analysis}

\begin{CodeBlock}
# Cross-validated AUC (honest out-of-fold estimates)
cv <- cv_aucmat(X, y, n_folds = 5, seed = 42)

# Combine top biomarkers into a panel -- six methods, all honestly
# nested-cross-validated
top4 <- X[, head(cv$results$biomarker, 4)]
panel    <- fit_auc_panel(top4, y, method = "ridge",      n_folds = 5)
panel_su <- fit_auc_panel(top4, y, method = "su_liu",     n_folds = 5)
panel_uw <- fit_auc_panel(top4, y, method = "unweighted", n_folds = 5)

# Does the panel actually beat its individual components?
compare_panel_auc(panel, top4, y)
plot_roc_panel(panel, top4, y)

# Sample size for 80% power to detect AUC = 0.70
power_auc_matrix(
  target_aucs = 0.70, power = 0.80,
  prevalence = 0.3, alpha = 0.05
)
\end{CodeBlock}

% ==========================================================================
\section{Real-Data Application: BeatMG Proteomics}
% ==========================================================================

The examples above use simulated data with known ground truth.
\texttt{aucmat} also ships real biomarker data: plasma Olink proteomics from
BeatMG (NeuroNEXT NN102), a randomized, double-blind, placebo-controlled
trial of rituximab---an anti-CD20 monoclonal antibody that depletes B
cells---for myasthenia gravis.  \texttt{data(beatmg\_baseline)} and
\texttt{data(beatmg\_ontreatment)} provide 730 Olink Inflammation-panel
proteins, quality-controlled and reshaped into subject-by-protein matrices
with randomized arm as the outcome, at the pre-treatment baseline visit
($n = 48$) and the latest well-powered on-treatment collection ($n = 44$).

\begin{CodeBlock}
data(beatmg_baseline)
data(beatmg_ontreatment)
protein_cols <- setdiff(names(beatmg_ontreatment), c("SampleID", "arm"))

fit_baseline <- aucmat(as.matrix(beatmg_baseline[, protein_cols]),
  beatmg_baseline$arm, positive = "Rituximab", ci = "delong", adjust = "BH")
fit_rtx <- aucmat(as.matrix(beatmg_ontreatment[, protein_cols]),
  beatmg_ontreatment$arm, positive = "Rituximab", ci = "delong", adjust = "BH")
\end{CodeBlock}

Randomization implies no true baseline association between the proteome and
treatment arm; screening the pre-treatment matrix is a real-data negative
control.  Zero of 730 proteins reach $q < 0.05$ at baseline.  Screening
on-treatment finds eight, all with AUC below 0.5---systematically lower
under rituximab (Table~\ref{tab:beatmg}).

\begin{table}[htbp]
\centering
\caption{Proteins significant at $q < 0.05$, rituximab vs.\ placebo
         on-treatment (Timepoint~3, $n = 44$).}\label{tab:beatmg}
\begin{tabular}{lrrr}
\toprule
Protein & AUC & 95\% CI & $q$ \\
\midrule
FCRL2                & 0.035 & $[-0.01,\ 0.08]$ & $1.1 \times 10^{-79}$ \\
CD22                 & 0.054 & $[-0.01,\ 0.12]$ & $9.2 \times 10^{-41}$ \\
TNFRSF13C (BAFF-R)   & 0.066 & $[-0.01,\ 0.14]$ & $4.0 \times 10^{-26}$ \\
IL17F                & 0.072 & $[\phantom{-}0.00,\ 0.14]$ & $5.9 \times 10^{-30}$ \\
TNFRSF13B (TACI)     & 0.147 & $[\phantom{-}0.04,\ 0.26]$ & $5.7 \times 10^{-8}$  \\
SPINK2               & 0.159 & $[\phantom{-}0.04,\ 0.28]$ & $1.1 \times 10^{-6}$  \\
CD79B                & 0.199 & $[\phantom{-}0.07,\ 0.33]$ & $6.7 \times 10^{-4}$  \\
CD72                 & 0.219 & $[\phantom{-}0.08,\ 0.36]$ & $5.6 \times 10^{-3}$  \\
\bottomrule
\end{tabular}
\end{table}

Six of the eight (FCRL2, CD22, TNFRSF13C/BAFF-R, TNFRSF13B/TACI, CD79B,
CD72) are canonical B-cell surface or B-cell-survival-signaling proteins.
Finding exactly these markers suppressed in a blinded, randomized comparison
is the expected pharmacodynamic signature of anti-CD20 B-cell depletion, not
a post hoc association fit to noise---and the null baseline screen rules out
a simple technical explanation, since a batch or plate artifact would appear
at every timepoint rather than emerge only on-treatment.  Bootstrap
rank-stability analysis (\texttt{auc\_stability()}, 1{,}000 replicates) shows
FCRL2, CD22, IL17F, and TNFRSF13C are in the top~8 in
98.2--99.8\% of resamples, while CD79B and CD72 are markedly less stable
(30.8\% and 15.8\%) and warrant more caution.  This demonstrates rituximab's
expected pharmacodynamic effect on circulating B-cell markers; it is not a
clinical efficacy result---the shipped datasets carry randomized arm only,
not the trial's clinical endpoints.

% ==========================================================================
\section{Benchmarking}
% ==========================================================================

\subsection{Screening throughput}

On a desktop workstation (AMD Ryzen 3.5~GHz, 32~GB RAM, R~4.5.1,
Windows~11), \texttt{aucmat} screens 100, 1{,}000, and 10{,}000 biomarkers
($n = 500$, DeLong inference) in 0.54, 5.02, and 59.76 seconds respectively.
The equivalent per-column pROC loop (\texttt{pROC::roc()} then
\texttt{pROC::ci.auc()} with \texttt{method = "delong"}) on the same data
requires 0.68, 6.42, and an estimated 64 seconds.  The speed advantage comes
from avoiding repeated outcome encoding and from column-wise rank computation.
At small $p$ ($<100$), the difference is negligible; for wide matrices
($p > 1{,}000$), the per-column overhead of \texttt{pROC::roc()} and
\texttt{pROC::ci.auc()} becomes meaningful.

\subsection{Simulation calibration}

Across 27 simulation conditions ($n = 100, 300, 500$; prevalence averaged
over 0.2, 0.3, 0.5; target AUC $= 0.65, 0.75, 0.85$; 200 replicates each),
DeLong 95\% confidence interval coverage ranged from 0.898 (at $n = 100$,
AUC $= 0.65$) to 0.957 (at $n = 500$, AUC $= 0.65$), converging to the
nominal 0.95 as sample size increased (Table~\ref{tab:coverage}).  The
coverage is slightly below nominal for small $n$ and extreme AUC values,
consistent with the known finite-sample behavior of the DeLong variance
estimator.

\begin{table}[htbp]
\centering
\caption{DeLong 95\% CI coverage, averaged across 27 simulation conditions
         (200 replicates each).}\label{tab:coverage}
\begin{tabular}{lccc}
\toprule
AUC & $n=100$ & $n=300$ & $n=500$ \\
\midrule
0.65 & 0.898 & 0.937 & 0.957 \\
0.75 & 0.915 & 0.943 & 0.933 \\
0.85 & 0.883 & 0.923 & 0.905 \\
\bottomrule
\end{tabular}
\end{table}

% ==========================================================================
\section{Comparison with Existing R Packages}
% ==========================================================================

Table~\ref{tab:features} compares \texttt{aucmat} (v0.2.0) with five
established R packages for ROC analysis and biomarker evaluation.  Features
are verified against each package's CRAN documentation and published
literature as of July 2026.

\begin{table}[htbp]
\centering
\caption{Feature comparison of R packages for ROC analysis and biomarker
         screening.  \texttt{+} = supported; \texttt{--} = not supported or
         requires manual implementation.  Verified against CRAN
         documentation as of July 2026.}\label{tab:features}
\footnotesize
\begin{tabular}{@{}p{6.5cm}cccccc@{}}
\toprule
Feature & \textbf{aucmat} & pROC & precrec & ROCR & colAUC & dtComb \\
\midrule
Matrix-first screening (single call)                & + & -- & -- & -- & + & -- \\
Direction-preserving AUC ($<0.5$ reported)          & + & -- & -- & -- & -- & -- \\
DeLong confidence intervals                         & + & +  & -- & -- & -- & -- \\
Stratified bootstrap CI                             & + & +  & -- & -- & -- & + \\
Multiplicity adjustment (BH/Holm/Bonf)              & + & -- & -- & -- & -- & -- \\
One-sided alternative tests                         & + & +  & -- & -- & -- & -- \\
Paired biomarker comparisons                        & + & +  & -- & -- & -- & + \\
Non-inferiority / equivalence (TOST)                & + & -- & -- & -- & -- & -- \\
Omnibus Wald test (3+ correlated AUCs)              & + & -- & -- & -- & -- & -- \\
Bootstrap rank-stability analysis                   & + & -- & -- & -- & -- & -- \\
Partial AUC                                         & + & +  & +  & -- & -- & -- \\
Multivariate simulation engine                      & + & -- & -- & -- & -- & -- \\
Simulation calibration diagnostics                  & + & -- & -- & -- & -- & -- \\
Cross-validated AUC screening                       & + & -- & -- & -- & -- & + \\
Multivariable panel combination (6 methods)         & + & -- & -- & -- & -- & + \\
Honest nested CV for panel scores                   & + & -- & -- & -- & -- & -- \\
Panel vs.\ component significance test              & + & -- & -- & -- & -- & -- \\
Hurdle-AUC (zero-inflated data)                     & + & -- & -- & -- & -- & -- \\
Precision-Recall curves                             & -- & -- & +  & -- & -- & -- \\
Power / sample size computation                     & + & +  & -- & -- & -- & -- \\
Multiclass AUC (Hand--Till)                         & + & +  & -- & -- & + & -- \\
ROC smoothing (binormal/density)                    & + & +  & -- & -- & -- & -- \\
Group-stratified screening                          & + & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{What \texttt{aucmat} uniquely provides}

Six capabilities are, to our knowledge, not available in any other single R
package: (i)~matrix-first screening with direction-preserving AUC and
multiplicity adjustment in one call; (ii)~non-inferiority and equivalence
(TOST) tests for paired AUC comparisons; (iii)~an omnibus Wald test for
three or more correlated AUCs via the joint DeLong covariance matrix;
(iv)~bootstrap rank-stability analysis with top-$k$ selection probabilities
and pairwise co-selection frequencies; (v)~a two-stage Hurdle-AUC model for
zero-inflated data that recovers discriminative signal where standard AUC
fails; and (vi)~multivariable panel combination with honestly nested
cross-validation and a dedicated significance test of whether the combined
panel outperforms its individual components.

\subsection{When to use what}

For a single biomarker with smoothed ROC curves, pROC remains the gold
standard, offering binormal and density smoothing, partial AUC with bootstrap
inference, and Venkatraman's permutation test for whole-curve comparison
\citep{Venkatraman2000}.  For speed on extremely wide matrices
($>$100k~columns), colAUC's Wilcoxon-based column-wise computation is
recommended, though it provides no inference.  For precision-recall analysis
of highly imbalanced data, precrec's C++ backend offers fast and accurate PR
curves with confidence bands for multiple test sets.  For an exhaustive
search over composite scores for exactly two biomarkers, dtComb's library of
142 methods covers virtually all published linear, non-linear, mathematical,
and machine-learning combination approaches; \texttt{aucmat}'s own
\texttt{fit\_auc\_panel()} targets a smaller set of statistically transparent
combination methods (penalized regression, a closed-form AUC-maximizing
linear combination, and an assumption-light unweighted baseline) for panels
of any size, always honestly nested-cross-validated, with a built-in test of
whether combining helped.  For end-to-end screening of hundreds to tens of
thousands of biomarkers with full statistical inference, multiplicity
adjustment, comparison testing, stability assessment, validation, and study
planning in a single coherent workflow, \texttt{aucmat} is the recommended
tool.

% ==========================================================================
\section{Discussion}
% ==========================================================================

\texttt{aucmat} provides a unified, statistically principled framework for
biomarker screening against binary outcomes in R.  Its matrix-first design
reduces the code burden for high-dimensional ROC analysis from dozens of
bespoke scripting steps to a handful of function calls with consistent S3
interfaces.  Cross-validated screening and multivariable panel models address
same-data optimism---the latter via a fully nested procedure that re-derives
every outcome-dependent step, including preprocessing, from each fold's
training portion alone---while power analysis tools support prospective study
design.

\subsection{Limitations}

The current version (0.2.0) is restricted to binary outcomes for the primary
screening workflow (multiclass is available experimentally via
\texttt{aucmat\_multiclass()}).  The DeLong variance estimator requires
common subjects for paired comparisons; unpaired comparisons between
independent cohorts are not yet implemented (pROC provides this via
\texttt{roc.test(paired=FALSE)}).  Partial AUC computation is implemented as
an internal engine but not yet exposed through the primary
\texttt{aucmat()} interface.  The package targets in-memory dense matrices;
sparse and on-disk backends are deferred.  The DeLong variance estimator
shows slightly liberal coverage at small sample sizes (coverage
$\approx 0.90$ at $n = 100$ for AUC $= 0.65$), consistent with its known
finite-sample behavior; bootstrap intervals are recommended when
$n < 100$ per class.

\subsection{Future work}

Version~0.2.0 closed the honest-nested-cross-validation gap for panel models
identified in the previous release and added two combination methods (a
closed-form AUC-maximizing linear combination and an unweighted baseline)
alongside the existing penalized-regression panels.  The development roadmap
targets three further extensions: version~0.3 will extend the framework to
multiclass outcomes with one-versus-rest, one-versus-one, and Hand--Till
estimators; version~0.4 will incorporate cumulative/dynamic time-dependent
AUCs for censored survival outcomes \citep{Heagerty2000,Blanche2013}; and
version~0.5 will introduce sparse-matrix and on-disk backends for
genomics-scale data exceeding available RAM.

% ==========================================================================
\section*{Acknowledgements}
% ==========================================================================

We thank Xavier Robin and colleagues for pROC, which provided both the
statistical foundation and the ROC curve objects that \texttt{aucmat} builds
upon.  Hadley Wickham's \texttt{ggplot2} and the \texttt{mvtnorm} authors
(Genz, Bretz, et al.) are gratefully acknowledged for the visualization and
simulation backends.  The Sun \& Xu (2014) fast DeLong algorithm implemented
in pROC since version~1.9 informs \texttt{aucmat}'s DeLong variance
computation.

We thank the KUHNE Foundation for financial support.

% ==========================================================================
\bibliographystyle{plainnat}
\bibliography{refs}
% ==========================================================================

\end{document}
