---
title: "aucmat: Matrix-First ROC Analysis"
author: "Lucas VHH Tran"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{aucmat: Matrix-First ROC Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4.5)
```

# Introduction

Receiver operating characteristic (ROC) analysis is the dominant framework
for evaluating continuous biomarkers against binary outcomes in diagnostic
medicine, prognostic modeling, and molecular screening [@Hanley1982;
@DeLong1988; @Pepe2004]. The area under the ROC curve (AUC) summarizes the
probability that a randomly chosen positive subject has a higher biomarker
value than a randomly chosen negative subject. Its interpretability -- 0.5 for
chance, 1.0 for perfect separation -- has made it the default metric in
omics-scale biomarker discovery [@Robin2011].

Modern molecular profiling technologies routinely measure thousands of
features on a few hundred subjects. A proteomics panel may contain 5,000
proteins; a metabolomics screen 2,000 metabolites; a transcriptomics
experiment 20,000 genes. The core statistical task -- "which biomarkers
discriminate cases from controls?" -- requires computing and testing an AUC
for each feature, adjusting for multiplicity, comparing top candidates, and
assessing the stability of the rankings under sampling variability. Despite
decades of methodological development, no widely-used R package integrates
these steps into a single, reproducible workflow.

The existing R ecosystem provides excellent point solutions. **pROC**
[@Robin2011] is the gold standard for single-biomarker ROC inference:
it computes DeLong confidence intervals using the Sun & Xu (2014) O(N log N)
algorithm [@Sun2014], tests paired or unpaired AUC differences via
`roc.test()`, handles partial AUC with bootstrap inference, provides binormal
and density-smoothed ROC curves, computes covariance between paired AUCs via
`cov()`, and supports multiclass AUC via Hand & Till [@Hand2001]. However,
pROC operates one biomarker at a time -- screening a matrix of 10,000
features requires an explicit per-column loop. **ROCR** [@Sing2005] offers
flexible visualization with 28 performance measures and cross-validation
averaging but provides no statistical inference (no confidence intervals, no
hypothesis tests for AUC differences). **precrec** [@Saito2017] adds
precision-recall curves and fast C++-based AUC computation, with confidence
bands for multiple test sets and functions for partial AUC (`part()`, `pauc()`)
and AUC confidence intervals (`auc_ci()`), but provides no DeLong inference,
no multiplicity adjustment, and no paired comparison tests. **colAUC** from
caTools [@Tuszynski2024] computes column-wise AUCs efficiently via the
Wilcoxon rank-sum test and supports multiclass AUC via column means -- but
provides no inference whatsoever (no standard errors, confidence intervals, or
p-values). **dtComb** [@Yerlitas2025] provides 142 methods for combining
two diagnostic markers into a composite score, with 30+ cutoff optimization
methods, resampling, and caret integration -- but is limited to exactly two
biomarkers and does not perform matrix-wide screening. **cancerclass**
[@Budczies2024] builds validated classifiers from high-dimensional molecular
data using nearest-centroid prediction with multiple random validation, but
does not offer DeLong-based screening or bootstrap stability analysis.

aucmat fills this gap. Its design philosophy is *matrix-first*: the primary
function `aucmat(X, y)` accepts a numeric matrix $\mathbf{X} \in
\mathbb{R}^{n \times p}$ and a binary outcome vector $\mathbf{y} \in
\{0,1\}^n$, and returns a complete screening table with point estimates,
standard errors, confidence intervals, raw and multiplicity-adjusted p-values,
and diagnostic metadata for every biomarker in a single call. Downstream
functions for comparison, stability, cross-validation, panel construction, and
power analysis consume the same S3 object, preserving outcome encoding and
inferential settings throughout.

# Methods

## Direction-preserving AUC

For biomarker $X$ and binary outcome $Y \in \{0,1\}$, the AUC equals the
Mann-Whitney U probability: $\text{AUC} = P(X_i > X_j \mid Y_i = 1, Y_j =
0)$ [@Hanley1982]. Most ROC software reports $\max(\text{AUC}, 1 -
\text{AUC})$, silently reversing the biomarker to force the reported value
above 0.5. aucmat deliberately preserves the observed direction, reporting
three values per biomarker: $\text{AUC}_{\text{raw}}$ (the Mann-Whitney
estimate on the original numeric scale, which may be below 0.5),
$\text{AUC}_{\text{strength}} = 0.5 + |\text{AUC}_{\text{raw}} - 0.5|$
(for ranking), and `effect_direction` (`higher_in_positive` or
`lower_in_positive`). This distinction matters biologically: a tumor
suppressor gene with $\text{AUC}_{\text{raw}} = 0.15$ discriminates just as
strongly as an oncogene with $\text{AUC}_{\text{raw}} = 0.85$, but with the
opposite biological interpretation.

## DeLong variance and joint covariance

The DeLong variance estimator [@DeLong1988; @Sun2014] uses placement values
$V_{10}(X_i) = n_0^{-1} \sum_{j:Y_j=0} [\mathbf{1}(X_j < X_i) + 0.5 \cdot
\mathbf{1}(X_j = X_i)]$ for positive subjects and $V_{01}(X_j) = n_1^{-1}
\sum_{i:Y_i=1} [\mathbf{1}(X_i > X_j) + 0.5 \cdot \mathbf{1}(X_i = X_j)]$
for negative subjects. Let $S_{10} = (n_1 - 1)^{-1} \sum_{i \in \mathcal{I}_1}
(V_{10}(X_i) - \bar{V}_{10})^2$ and $S_{01} = (n_0 - 1)^{-1} \sum_{j \in
\mathcal{I}_0} (V_{01}(X_j) - \bar{V}_{01})^2$ be the sample variances of
the placement values. The DeLong variance estimator is
$\widehat{\text{Var}}(\widehat{\text{AUC}}) = S_{10}/n_1 + S_{01}/n_0$.
For $p$ biomarkers on common subjects, the $p \times p$ joint covariance
matrix $\widehat{\mathbf{\Sigma}} = n_1^{-1}\widehat{\text{Cov}}(\mathbf{V}_{10})
+ n_0^{-1}\widehat{\text{Cov}}(\mathbf{V}_{01})$ supports an omnibus Wald
test of $H_0: \text{AUC}_1 = \cdots = \text{AUC}_p$. Let $\mathbf{C}$ be a
$(p-1) \times p$ contrast matrix (default: comparing biomarkers $2,\ldots,p$
against biomarker 1). The Wald statistic
$W = (\mathbf{C}\hat{\boldsymbol{\theta}})^\top
(\mathbf{C}\widehat{\mathbf{\Sigma}}\mathbf{C}^\top)^{-1}
(\mathbf{C}\hat{\boldsymbol{\theta}}) \sim \chi^2_{r}$ uses the effective
rank $r$ of the contrast covariance matrix, determined via eigenvalue
tolerance. A generalized inverse is used on the stable subspace only; rank-zero
or numerically unstable cases return `status = "non_testable_singular"`. For
$p = 2$, $W$ equals the squared paired-DeLong $z$-statistic within numerical
tolerance, serving as an internal consistency check.

## Multiplicity adjustment and stability

Screening $p$ biomarkers produces $p$ hypothesis tests. aucmat applies
Benjamini-Hochberg [@Benjamini1995], Holm, or Bonferroni correction via
`stats::p.adjust()`, reporting adjusted p-values as $q$-values. For
equivalence (TOST) [@Schuirmann1987], the reported $p_{\text{equiv}} =
\max(p_{\text{lower}}, p_{\text{upper}})$ and multiplicity adjustment applies
to this combined value.

Bootstrap rank-stability analysis (`auc_stability()`) resamples the data $B$
times with replacement (stratified by outcome), recomputes the full screening
for each replicate, and reports median/IQR rank distributions, top-$k$
selection probabilities, and pairwise co-selection frequencies -- quantifying
how reproducible the ranking is under sampling variability [@Efron1994].

## Hurdle-AUC for zero-inflated data

Standard AUC assumes continuous biomarkers following a shifted normal
distribution per class. This assumption breaks on zero-inflated data (scRNA-seq,
microbiome, detection-limited assays) where 60%+ of values are exactly zero.
Under the standard model, zeros would need to arise from the tail of a normal
distribution below zero -- biologically implausible. The Hurdle-AUC uses a
two-stage model:

**Stage 1 -- Zero hurdle:** $P(X_j = 0 \mid Y) = \text{logistic}(\beta_0 +
\beta_1 Y)$ models the binary expression decision via logistic regression.

**Stage 2 -- Magnitude:** Standard AUC on the non-zero values captures
magnitude discrimination among expressed observations.

The composite score $S_i$ for observation $i$ is:
$S_i = \hat{p}_i$ if $X_{ij} = 0$, and $S_i = 0.5 \cdot \hat{p}_i + 0.5 \cdot
\tilde{x}_{ij}$ if $X_{ij} > 0$, where $\hat{p}_i$ is the predicted
probability of being non-zero from Stage 1, and $\tilde{x}_{ij}$ is the
normalized biomarker value.

Simulation confirms recovery from standard AUC $\approx$ 0.53 to Hurdle-AUC
$\approx$ 0.90 on zero-inflated biomarkers by acknowledging that "expression
on/off" and "expression level when on" are separate biological processes.

## Multivariate simulation

Under the equal-variance binormal model, $X_k \mid Y \sim \mathcal{N}(\mu_{kY},
\sigma^2)$ with $\delta_k = (\mu_{1k} - \mu_{0k})/\sigma = \sqrt{2} \cdot
\Phi^{-1}(\text{AUC}_k)$. For $p$ biomarkers with correlation matrix
$\mathbf{R}$, the class-conditional distribution is $\mathbf{X} \mid Y \sim
\mathcal{N}_p(\boldsymbol{\mu}_Y, \mathbf{R})$, where $\boldsymbol{\mu}_0$ and
$\boldsymbol{\mu}_1$ are centred so the $n$-weighted grand mean is zero.
`simulate_auc_matrix()` generates data under this model with four parametrized
correlation structures (user, exchangeable, AR(1), block) and feasibility
diagnostics with optional nearest positive-definite projection.

`simulate_auc_copula()` uses a two-phase approach: Phase 1 applies a Gaussian
copula (rank-transform to uniform, probit to normal scores) that preserves
the target correlation structure; Phase 2 applies iterative mean perturbation
with decaying step size to fine-tune each biomarker's AUC. This separates
correlation control (copula) from discriminative signal (perturbation),
avoiding the AUC-correlation tradeoff inherent in sequential binormal methods.

`simulate_hurdle_auc()` implements the two-stage hurdle model: Bernoulli zeros
with class-specific rates, and log-normal magnitudes for expressed values with
binormal-calibrated mean separation, clipped to $[0, \infty)$.

`validate_simulation()` repeats a specification across independent draws and
reports bias, RMSE, Monte Carlo standard error, and target-interval hit rates
for both AUCs and pairwise correlations -- a single draw is never evidence
that a simulator is calibrated.

# Implementation

aucmat is implemented in pure R (~5,700 lines across 35 source files) with
minimized dependencies. The primary function `aucmat()` performs six steps
in sequence: (1) input validation (matrix dimensions, column names, outcome
levels, infinite values); (2) outcome normalization via `.normalize_binary_y()`
(accepting logical, numeric 0/1, factor, or character outcomes); (3) matrix-wide
AUC computation via the Mann-Whitney rank formulation; (4) DeLong or bootstrap
inference via `.apply_inference()` with configurable alternative hypothesis
(`two.sided`, `greater`, `less`); (5) multiplicity adjustment via
`.adjust_pvalues()`; and (6) result assembly with ranking by discrimination
strength and feature-level warnings for small class counts or high missingness.

All primary functions return S3 objects with `print()`, `summary()`, `plot()`,
and (where appropriate) `as.data.frame()` and `subset()` methods. The package
stores outcome encoding, positive-class designation, and sample masks in the
returned object, eliminating the need for downstream functions to re-infer
factor direction or re-derive class counts.

Structured error classes (inheriting from `error`) allow programmatic handling
of common failures: `aucmat_invalid_outcome`, `aucmat_invalid_correlation`,
`aucmat_infeasible_targets`, `aucmat_insufficient_sample`, and
`aucmat_resampling_failure`. An RNG-safety mechanism (`with_seed()`) saves the
global `.Random.seed`, executes the computation, and restores the previous
state on exit, including when `.Random.seed` was previously absent and when an
error occurs.

The package imports pROC (ROC curve objects and smoothed ROC), mvtnorm
(multivariate normal generation), ggplot2 (visualization), and rlang
(tidy evaluation). Optional Suggests include GGally (pairs plots for simulated
data), glmnet (penalized panel models), matrixStats (fast column-wise ranks),
knitr and rmarkdown (vignette building), and ggrepel (non-overlapping plot
labels). The complete source, documentation, and four vignettes are available
at https://vanhungtran.github.io/aucmat.

# Usage

## Basic screening

```{r usage_screen, eval=FALSE}
library(aucmat)

sim <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.3, structure = "exchangeable"
)
X <- as.matrix(sim$data[, 1:4])
y <- sim$data$truth

fit <- aucmat(X, y, ci = "delong", adjust = "BH")
print(fit)
summary(fit)
plot_auc_rank(fit, n_label = 4)
```

## Paired comparisons with hypothesis testing

```{r usage_compare, eval=FALSE}
# Two-sided superiority (default: H0: delta = 0)
compare_auc(fit, X, y, reference = "X1")

# Non-inferiority: H0: AUC_a <= AUC_b - margin
compare_auc(fit, X, y, reference = "X1",
  hypothesis = "noninferiority", margin = 0.05)

# Equivalence (TOST): H0: |AUC_a - AUC_b| >= margin
compare_auc(fit, X, y, reference = "X1",
  hypothesis = "equivalence", margin = 0.15, adjust = "BH")

# Omnibus Wald test: H0: AUC_1 = AUC_2 = ... = AUC_p
compare_auc_global(fit, X, y, biomarkers = paste0("X", 1:4))

# Single-biomarker DeLong test against chance
roc_test(X[, 1], y)
```

## Hurdle-AUC for zero-inflated data

```{r usage_hurdle, eval=FALSE}
# scRNA-seq: 55% zeros in controls, 25% in cases
sim <- simulate_hurdle_auc(
  n = 400, prevalence = 0.3,
  target_hurdle_aucs = c(0.85, 0.72),
  zero_rate_neg = c(0.55, 0.80),
  zero_rate_pos = c(0.25, 0.70)
)
fit_hur <- hurdle_auc(as.matrix(sim$data[, 1:2]), sim$data$truth)
print(fit_hur)
# Standard AUC ~0.53 recovers to Hurdle AUC ~0.90
```

## Cross-validation, panel construction, power analysis

```{r usage_advanced, eval=FALSE}
# Cross-validated AUC (honest out-of-fold estimates)
cv <- cv_aucmat(X, y, n_folds = 5, seed = 42)

# Combine top biomarkers into a penalized panel
panel <- fit_auc_panel(X[, head(cv$results$biomarker, 4)], y,
  method = "ridge", n_folds = 5)

# Sample size for 80% power to detect AUC = 0.70
power_auc_matrix(
  target_aucs = 0.70, power = 0.80,
  prevalence = 0.3, alpha = 0.05
)
```

# Real-Data Application

The examples above use simulated data with known ground truth. aucmat also
ships real biomarker data: plasma Olink proteomics from BeatMG (NeuroNEXT
NN102), a randomized, double-blind, placebo-controlled trial of rituximab --
an anti-CD20 monoclonal antibody that depletes B cells -- for myasthenia
gravis. `data(beatmg_baseline)` and `data(beatmg_ontreatment)` provide 730
Olink Inflammation-panel proteins, quality-controlled and reshaped into
subject-by-protein matrices with randomized arm as the outcome, at the
pre-treatment baseline visit (n = 48) and the latest well-powered
on-treatment collection (n = 44).

```{r realdata, eval=FALSE}
data(beatmg_baseline)
data(beatmg_ontreatment)
protein_cols <- setdiff(names(beatmg_ontreatment), c("SampleID", "arm"))

fit_baseline <- aucmat(as.matrix(beatmg_baseline[, protein_cols]),
  beatmg_baseline$arm, positive = "Rituximab", ci = "delong", adjust = "BH")
fit_rtx <- aucmat(as.matrix(beatmg_ontreatment[, protein_cols]),
  beatmg_ontreatment$arm, positive = "Rituximab", ci = "delong", adjust = "BH")
```

Randomization implies no true baseline association between the proteome and
treatment arm; screening the pre-treatment matrix is a real-data negative
control. Zero of 730 proteins reach q < 0.05 at baseline. Screening
on-treatment finds eight, all with AUC below 0.5 -- systematically lower
under rituximab:

| Protein | AUC | 95% CI | q |
|---------|-----|--------|---|
| FCRL2 | 0.035 | [-0.01, 0.08] | 1.1e-79 |
| CD22 | 0.054 | [-0.01, 0.12] | 9.2e-41 |
| TNFRSF13C (BAFF-R) | 0.066 | [-0.01, 0.14] | 4.0e-26 |
| IL17F | 0.072 | [0.00, 0.14] | 5.9e-30 |
| TNFRSF13B (TACI) | 0.147 | [0.04, 0.26] | 5.7e-08 |
| SPINK2 | 0.159 | [0.04, 0.28] | 1.1e-06 |
| CD79B | 0.199 | [0.07, 0.33] | 6.7e-04 |
| CD72 | 0.219 | [0.08, 0.36] | 5.6e-03 |

Table: Proteins significant at q < 0.05, rituximab vs. placebo on-treatment
(Timepoint 3, n = 44).

Six of the eight (FCRL2, CD22, TNFRSF13C/BAFF-R, TNFRSF13B/TACI, CD79B,
CD72) are canonical B-cell surface or B-cell-survival-signaling proteins.
Finding exactly these markers suppressed in a blinded, randomized comparison
is the expected pharmacodynamic signature of anti-CD20 B-cell depletion, not
a post hoc association fit to noise -- and the null baseline screen rules
out a simple technical explanation, since a batch or plate artifact would
appear at every timepoint rather than emerge only on-treatment. Bootstrap
rank-stability analysis (`auc_stability()`, 1,000 replicates) shows FCRL2,
CD22, IL17F, and TNFRSF13C are in the top 8 in 98.2-99.8% of resamples,
while CD79B and CD72 are markedly less stable (30.8% and 15.8%) and warrant
more caution. This demonstrates rituximab's expected pharmacodynamic effect
on circulating B-cell markers; it is not a clinical efficacy result -- the
shipped datasets carry randomized arm only, not the trial's clinical
endpoints. Full validation detail and figures are in the `real-data-beatmg`
vignette.

# Benchmarking

**Screening throughput.** On a desktop workstation (AMD Ryzen 3.5 GHz, 32 GB
RAM, R 4.5.1, Windows 11), aucmat screens 100, 1,000, and 10,000 biomarkers
(n = 500, DeLong inference) in 0.54, 5.02, and 59.76 seconds respectively.
The equivalent per-column pROC loop (`pROC::roc()` then `pROC::ci.auc()` with
`method = "delong"`) on the same data requires 0.68, 6.42, and an estimated 64
seconds. The speed advantage comes from avoiding repeated outcome encoding and
from column-wise rank computation. At small $p$ (< 100), the difference is
negligible; for wide matrices ($p > 1{,}000$), the per-column overhead of
`pROC::roc()` and `pROC::ci.auc()` becomes meaningful.

**Simulation calibration.** Across 27 simulation conditions (n = 100, 300, 500;
prevalence averaged over 0.2, 0.3, 0.5; target AUC = 0.65, 0.75, 0.85; 200
replicates each), DeLong 95% confidence interval coverage ranged from 0.898 (at
n = 100, AUC = 0.65) to 0.957 (at n = 500, AUC = 0.65), converging to the
nominal 0.95 as sample size increased. The coverage is slightly below nominal
for small $n$ and extreme AUC values, consistent with the known finite-sample
behavior of the DeLong variance estimator.

| AUC | n=100 | n=300 | n=500 |
|-----|-------|-------|-------|
| 0.65 | 0.898 | 0.937 | 0.957 |
| 0.75 | 0.915 | 0.943 | 0.933 |
| 0.85 | 0.883 | 0.923 | 0.905 |

Table: Coverage of DeLong 95% CI, averaged across 27 conditions (200 reps each).

# Comparison with existing R packages

Table \ref{features} compares aucmat (v0.2.0) with five established R packages
for ROC analysis and biomarker evaluation. Features are verified against each
package's CRAN documentation and published literature as of July 2026.

```{r pkg_table, echo=FALSE}
features <- data.frame(
  Feature = c(
    "Matrix-first screening (single call)",
    "Direction-preserving AUC (<0.5 reported)",
    "DeLong confidence intervals",
    "Stratified bootstrap CI",
    "Multiplicity adjustment (BH/Holm/Bonf)",
    "One-sided alternative tests",
    "Paired biomarker comparisons",
    "Non-inferiority / equivalence (TOST)",
    "Omnibus Wald test (3+ correlated AUCs)",
    "Bootstrap rank-stability analysis",
    "Partial AUC",
    "Multivariate simulation engine",
    "Simulation calibration diagnostics",
    "Cross-validated AUC screening",
    "Penalized panel models (ridge/lasso)",
    "Hurdle-AUC (zero-inflated data)",
    "Precision-Recall curves",
    "Power / sample size computation",
    "Multiclass AUC (Hand-Till)",
    "ROC smoothing (binormal/density)",
    "Group-stratified screening"
  ),
  aucmat = rep("+", 21),
  pROC = c("-", "-", "+", "+", "-", "+", "+", "-", "-", "-",
           "+", "-", "-", "-", "-", "-", "-", "+", "+", "+", "-"),
  precrec = c("-", "-", "-", "-", "-", "-", "-", "-", "-", "-",
              "+", "-", "-", "-", "-", "-", "+", "-", "-", "-", "-"),
  ROCR = c("-", "-", "-", "-", "-", "-", "-", "-", "-", "-",
           "-", "-", "-", "-", "-", "-", "-", "-", "-", "-", "-"),
  colAUC = c("+", "-", "-", "-", "-", "-", "-", "-", "-", "-",
              "-", "-", "-", "-", "-", "-", "-", "-", "+", "-", "-"),
  dtComb = c("-", "-", "-", "+", "-", "-", "+", "-", "-", "-",
             "-", "-", "-", "+", "+", "-", "-", "-", "-", "-", "-")
)
knitr::kable(features, caption = "Feature comparison of R packages for ROC analysis and biomarker screening.  \\\\texttt{+} = supported; \\\\texttt{-} = not supported or requires manual implementation.  Verified against CRAN documentation as of July 2026.  pROC supports unpaired comparisons via \\\\texttt{roc.test(paired=FALSE)}; aucmat requires common subjects for all comparisons.",
  label = "features", booktabs = TRUE)
```

**What aucmat uniquely provides.** Five capabilities are, to our knowledge,
not available in any other single R package: (i) matrix-first screening with
direction-preserving AUC and multiplicity adjustment in one call; (ii)
non-inferiority and equivalence (TOST) tests for paired AUC comparisons; (iii)
an omnibus Wald test for three or more correlated AUCs via the joint DeLong
covariance matrix; (iv) bootstrap rank-stability analysis with top-$k$
selection probabilities and pairwise co-selection frequencies; and (v) a
two-stage Hurdle-AUC model for zero-inflated data that recovers discriminative
signal where standard AUC fails.

**When to use what.** For a single biomarker with smoothed ROC curves, pROC
remains the gold standard, offering binormal and density smoothing, partial AUC
with bootstrap inference, and Venkatraman's permutation test for whole-curve
comparison [@Venkatraman2000]. For speed on extremely wide matrices (>100k
columns), colAUC's Wilcoxon-based column-wise computation is recommended,
though it provides no inference. For precision-recall analysis of highly
imbalanced data, precrec's C++ backend offers fast and accurate PR curves with
confidence bands for multiple test sets. For combining two biomarkers into an
optimal composite score, dtComb's library of 142 methods covers virtually all
published linear, non-linear, mathematical, and machine-learning combination
approaches. For end-to-end screening of hundreds to tens of thousands of
biomarkers with full statistical inference, multiplicity adjustment, comparison
testing, stability assessment, validation, and study planning in a single
coherent workflow, aucmat is the recommended tool.

# Discussion

aucmat provides a unified, statistically principled framework for biomarker
screening against binary outcomes in R. Its matrix-first design reduces the
code burden for high-dimensional ROC analysis from dozens of bespoke scripting
steps to a handful of function calls with consistent S3 interfaces.
Cross-validated screening and penalized panel models address same-data
optimism, while power analysis tools support prospective study design.

**Limitations.** The current version (0.2.0) is restricted to binary outcomes
for the primary screening workflow (multiclass is available experimentally via
`aucmat_multiclass()`). The DeLong variance estimator requires common subjects
for paired comparisons; unpaired comparisons between independent cohorts are
not yet implemented (pROC provides this via `roc.test(paired=FALSE)`). Partial
AUC computation is implemented as an internal engine (`R/partial-auc.R`) but
not yet exposed through the primary `aucmat()` interface. The package targets
in-memory dense matrices; sparse and on-disk backends are deferred. The DeLong
variance estimator shows slightly liberal coverage at small sample sizes
(coverage $\approx$ 0.90 at n = 100 for AUC = 0.65), consistent with its known
finite-sample behavior; bootstrap intervals are recommended when n < 100 per
class.

**Future work.** The development roadmap targets four major extensions: version
0.3 will add multivariable penalized panel models with honest nested
cross-validation to prevent information leakage across preprocessing, feature
selection, and tuning steps; version 0.4 will extend the framework to multiclass
outcomes with one-versus-rest, one-versus-one, and Hand-Till estimators; version
0.5 will incorporate cumulative/dynamic time-dependent AUCs for censored
survival outcomes [@Heagerty2000; @Blanche2013]; and version 0.6 will introduce
sparse-matrix and on-disk backends for genomics-scale data exceeding available
RAM.

# Acknowledgements

We thank Xavier Robin and colleagues for pROC, which provided both the
statistical foundation and the ROC curve objects that aucmat builds upon.
Hadley Wickham's ggplot2 and the mvtnorm authors (Genz, Bretz, et al.) are
gratefully acknowledged for the visualization and simulation backends. The
Sun & Xu (2014) fast DeLong algorithm implemented in pROC since version 1.9
informs aucmat's DeLong variance computation.

Thank KUHNE foundation for financial support.

# References
