---
title: "Introduction to aucmat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aucmat)
```

## Generating Simulated Data

The `aucmat` package provides a function called `generate_data_analytical` to create synthetic datasets with known properties. This is useful for testing and validating analysis methods.

In this example, we'll generate a dataset with 500 samples, a 30% prevalence of the positive class, and three biomarkers with target AUC values of 0.9, 0.8, and 0.7. We'll also specify a correlation matrix for the biomarkers.

```{r generate_data}
# Define the target AUCs and correlation matrix
target_aucs <- c(0.9, 0.8, 0.7)
corr_matrix <- matrix(c(
  1.0, 0.3, 0.1,
  0.3, 1.0, 0.2,
  0.1, 0.2, 1.0
), nrow = 3)

# Generate the synthetic data
sim_data <- generate_data_analytical(
  n = 5000,
  prevalence = 0.3,
  target_aucs = target_aucs,
  corr_matrix = corr_matrix
)

# Inspect the generated data
head(sim_data$data)
```

## Analyzing the Simulated Data

Now that we have a simulated dataset, we can use the `tableroc` function to analyze the biomarkers and see how well they perform.

```{r analyze_data}
# Extract the data and outcome
# The generated data has predictor columns `X1..Xp` and a final `truth` column.
X <- subset(sim_data$data, select = -truth) # Predictors
y <- sim_data$data$truth  # Outcome

# Run the tableroc function
results <- tableroc(X, y)

# Print the results
print(results)
```

## Plotting ROC curves for predictor combinations

The `plot_roc_with_combos()` helper builds ROC curves for all combinations
of predictors (or a restricted set via `combo_sizes`) and returns a **list**
containing:

- `plot`: the ggplot2 object with ROC curves and legend
- `roc_list`: list of pROC ROC objects for each combination
- `auc`: vector of AUC values
- `auc_sd`: vector of AUC standard deviations (when `add_ci = TRUE`)
- `ci_df`: data frame with confidence interval data
- `skipped_ci`: names of curves that skipped CI computation (e.g., perfect AUC)

Below we call it on the simulated dataset and render the plot. Note that we
access the plot via `res_plot$plot` and the legend automatically displays
on the right side, showing each model with its AUC ± SD.

```{r plot_combos, fig.width=6, fig.height=6}
# Build predictors vector (all X* columns)
preds <- setdiff(names(sim_data$data), "truth")

# Create the plot for all single and multi-variable combinations
res_plot <- plot_roc_with_combos(
  data = sim_data$data,
  outcome = "truth",
  predictors = preds,
  combo_sizes = 1:3,
  add_ci = TRUE,
  conf_level = 0.95,
  boot_n = 500,      # smaller boot for vignette speed
  seed = 123,
  auc_digits = 4,
   sd_digits = 4, 
  show_auc_labels = TRUE,
  title = "ROC curves for predictor combinations (simulated data)"
)

# Print ggplot
print(res_plot$plot)

# Show AUCs for reference
res_plot$auc
```

## Correlation heatmap of predictors

It's often useful to inspect pairwise correlations between predictors. The
chunk below computes the correlation matrix for the predictor columns and
renders a heatmap with values.

```{r corr_heatmap, fig.width=6, fig.height=5}
# Predictor names
preds <- setdiff(names(sim_data$data), "truth")

# Compute correlation matrix (pairwise complete observations)
cor_mat <- stats::cor(sim_data$data[, preds], use = "pairwise.complete.obs")
cor_df <- as.data.frame(as.table(cor_mat))

# Plot heatmap with ggplot2
p_cor <- ggplot2::ggplot(cor_df, ggplot2::aes(Var1, Var2, fill = Freq)) +
  ggplot2::geom_tile() +
  ggplot2::scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  ggplot2::geom_text(ggplot2::aes(label = sprintf("%.2f", Freq)), size = 3) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::labs(title = "Correlation heatmap of predictors", x = NULL, y = NULL)

print(p_cor)
```