---
title: "Introduction to aucmat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{Introduction to aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r setup}
library(aucmat)
```

## Overview

`aucmat` is a matrix-first R package for screening and ranking molecular
biomarkers against binary outcomes. It reports **direction-preserving AUCs**
and provides inference, multiplicity adjustment, paired comparisons, bootstrap
rank stability, and publication-quality visualizations.

---

## 1. How Data is Simulated (The Simple Version)

The AUC (Area Under the ROC Curve) answers: *if I pick one positive and one
negative subject at random, how often does the positive have a higher
biomarker value?*

An AUC of 0.5 means the biomarker is useless (coin flip). An AUC of 1.0
means perfect separation.

### The key idea

Under the **binormal model**, biomarker values are normally distributed in
each class, shifted apart:

```
Negative class:  X | Y=0  ~  N(0, 1)
Positive class:  X | Y=1  ~  N(delta, 1)
```

The bigger the shift `delta`, the higher the AUC:

$$AUC = \Phi\left(\frac{\delta}{\sqrt{2}}\right)$$

where $\Phi$ is the standard normal CDF. To get a target AUC, we invert this:

$$\delta = \sqrt{2} \cdot \Phi^{-1}(AUC)$$

For example, AUC = 0.8 requires $\delta \approx 1.19$ — the positive class
mean is about 1.19 standard deviations above the negative class mean.

### Adding correlation between biomarkers

`aucmat` provides three approaches:

- **`simulate_auc_matrix()`**: biomarkers are drawn class-conditionally
  (`X | Y ~ MVN(mu_class, R)`) with a named correlation `structure`
  (`exchangeable`, `ar1`, `block`, or a hand-supplied `user` matrix) instead
  of a full latent matrix, plus feasibility diagnostics. Fast (closed-form
  calibration, no numerical search) and the recommended starting point for
  most use cases — see the
  [Simulating Biomarker Data](simulating-biomarker-data.html) vignette.

- **`generate_data_probit()`**: all biomarkers + a latent outcome variable
  come from one big multivariate normal distribution. AUCs and correlations
  are both free parameters, at the cost of numerical calibration.

- **`generate_data_analytical()`**: biomarkers are built one at a time from
  highest to lowest AUC, each inheriting correlation from previous ones
  through sequential regression. Fast but AUC and correlation are linked.

---

## 2. Simulating Biomarker Data

### 2.1 Latent Probit (AUC + Correlation Independent)

```{r sim_probit}
set.seed(42)
sim <- generate_data_probit(
  n = 500,
  target_aucs = c(0.90, 0.80, 0.70, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.4, 0.2, 0.1,
    0.4, 1.0, 0.3, 0.2,
    0.2, 0.3, 1.0, 0.3,
    0.1, 0.2, 0.3, 1.0
  ), 4, 4),
  prevalence = 0.3
)

# Target vs achieved AUCs
data.frame(Target = sim$target_aucs,
           Achieved = round(sim$achieved_aucs, 4))
```

### 2.2 Binormal Sequential (Fast)

```{r sim_binormal}
sim2 <- generate_data_analytical(
  n = 500,
  prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  corr_matrix = matrix(c(1, 0.3, 0.1, 0.3, 1, 0.2, 0.1, 0.2, 1), 3, 3)
)
round(sim2$achieved_aucs, 4)
```

### 2.3 Single AUC Vector (Exact Empirical AUC)

```{r sim_auc_vec}
y <- rbinom(200, 1, 0.3)
out <- generate_auc_vector(y, target_auc = 0.80)
c(Achieved_AUC = out$achieved_auc)
```

### 2.4 AUC + Correlation for a Single Biomarker

```{r sim_auc_cor}
out2 <- generate_auc_cor_vector(y, target_auc = 0.80, target_cor = 0.45)
c(AUC = out2$achieved_auc, Correlation = out2$achieved_cor)
```

### 2.5 Monte Carlo Sampling Distribution

```{r sim_monte}
sim_mc <- simulate_auc_correlation(y, target_auc = c(0.7, 0.8, 0.9),
                                    n_sim = 200, seed = 123)
head(sim_mc$results, 6)
```

### 2.6 General-Purpose Matrix Simulator (Recommended)

```{r sim_matrix}
sim_matrix <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.3, structure = "exchangeable", seed = 1
)
data.frame(target = sim_matrix$target_aucs,
           achieved = round(sim_matrix$achieved_aucs, 4))
```

`structure` also accepts `"ar1"`, `"block"`, or a hand-supplied `"user"`
matrix, and `validate_simulation()` repeats a spec across many draws to
confirm it's calibrated (bias, RMSE, hit rate) — see the
[Simulating Biomarker Data](simulating-biomarker-data.html) vignette for
details.

---

## 3. Matrix Screening

`aucmat()` screens every biomarker column against a binary outcome.

```{r screen}
X <- as.matrix(sim$data[, 1:4])
y <- sim$data$truth

fit <- aucmat(X, y, ci = "delong", adjust = "BH")
print(fit)
```

```{r summary}
summary(fit)
```

### Result columns

| Column | Description |
|--------|-------------|
| `auc_raw` | AUC on the observed direction (may be < 0.5) |
| `auc_strength` | $0.5 + \|\text{auc\_raw} - 0.5\|$ — always ≥ 0.5 |
| `effect_direction` | `higher_in_positive` or `lower_in_positive` |
| `std_error` | DeLong or bootstrap standard error |
| `conf_low`, `conf_high` | Confidence interval |
| `p_value`, `q_value` | Raw and multiplicity-adjusted |
| `rank` | By `auc_strength` descending |
| `status` | `ok`, `constant`, `insufficient_positive`, etc. |

---

## 4. Visualizing Results

### 4.1 Rank Plot — Ordered Discrimination Strengths

```{r plot_rank, fig.width=7, fig.height=4}
plot_auc_rank(fit, n_label = 4)
```

### 4.2 Volcano Plot — Effect vs Evidence

```{r plot_volcano, fig.width=7, fig.height=4}
plot_auc_volcano(fit, q_cutoff = 0.05)
```

### 4.3 Forest Plot — Multiple AUCs with Confidence Intervals

```{r plot_forest, fig.width=7, fig.height=3.5}
plot_auc_forest(fit, n = 4)
```

### 4.4 ROC Curves — Multiple Biomarkers Overlaid

```{r plot_roc, fig.width=5.5, fig.height=5}
plot_roc_top(fit, X = X, y = y, biomarkers = head(fit$results$biomarker, 3))
```

### 4.5 Stability Plot — Bootstrap Rank Distributions

```{r plot_stability, fig.width=8, fig.height=4}
stab <- auc_stability(X, y, times = 200, top_k = c(3, 5), seed = 42)
plot_auc_stability(stab, n_label = 8)
```

---

## 5. Comparing Biomarker AUCs

`compare_auc()` provides paired comparisons using DeLong's test on common
subjects. **Different comparison strategies**:

### 5.1 Compare All Pairs Among Top Biomarkers

```{r compare_top}
cmp_top <- compare_auc(fit, X, y, top_n = 3)
print(cmp_top)
```

### 5.2 Compare All Others to a Reference Biomarker

```{r compare_ref}
cmp_ref <- compare_auc(fit, X, y, reference = "X1")
print(cmp_ref)
```

### 5.3 Compare a Specific Set of Biomarkers

```{r compare_set}
cmp_set <- compare_auc(fit, X, y, biomarkers = c("X1", "X2", "X4"))
print(cmp_set)
```

### 5.4 Adjust for Multiplicity in Comparisons

```{r compare_adj}
cmp_adj <- compare_auc(fit, X, y, top_n = 3, adjust = "BH")
print(cmp_adj)
```

The output includes: `auc_a`, `auc_b`, `auc_diff`, `std_error`, `conf_low`,
`conf_high`, `p_value`, `q_value`, and common sample counts per pair.

### 5.5 Global Test: Are All AUCs Equal?

`compare_auc_global()` complements the pairwise comparisons above with an
omnibus Wald test of H0: AUC_1 = AUC_2 = ... = AUC_p on a common subject
set, using the joint DeLong covariance matrix:

```{r compare_global}
global_test <- compare_auc_global(fit, X, y)
print(global_test)
```

---

## 6. Rank Stability

Bootstrap resampling quantifies how much rankings change under sampling variability.

```{r stability}
stab <- auc_stability(X, y, times = 500, top_k = c(10, 25), seed = 42)
print(stab)

# Top-10 selection probabilities
head(stab$top_k_probs[stab$top_k_probs$k == 10, ], 6)
```

```{r stability_plot, fig.width=8, fig.height=4}
plot_auc_stability(stab, n_label = 10)
```

---

## 7. Handling Missing Data

Two strategies via `na_action`:

- **`featurewise`** (default): each biomarker uses all subjects with observed
  values — maximises available data
- **`complete`**: only subjects with complete observations across all
  biomarkers — ensures comparable populations

```{r missing}
X_na <- X
X_na[sample(length(X_na), 50)] <- NA

fit_na <- aucmat(X_na, y, ci = "none", na_action = "featurewise")
head(fit_na$results[, c("biomarker", "n_used", "n_missing", "missing_fraction")])
```

Imputation should be done **before** calling `aucmat()`. Outcome-dependent
imputation is not performed internally.

---

## 8. Reproducibility

All functions accept a `seed` argument:

```{r repro}
fit1 <- aucmat(X, y, ci = "bootstrap", boot_n = 500, seed = 123)
fit2 <- aucmat(X, y, ci = "bootstrap", boot_n = 500, seed = 123)
identical(fit1$results$auc_raw, fit2$results$auc_raw)
```

---

## Getting Help

```r
?aucmat
?simulate_auc_matrix
?validate_simulation
?generate_data_probit
?compare_auc
?compare_auc_global
?auc_stability
?plot_auc_rank
```
