---
title: "Introduction to aucmat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

```{r setup}
library(aucmat)
```

## Overview

`aucmat` is a matrix-first R package for scalable and statistically principled
ROC analysis of high-dimensional molecular biomarker data.  Its primary task is
**univariate screening and ranking** of genes, proteins, and metabolites
measured on the same subjects.

Key features:

- **Direction-preserving AUC** — never silently reverses biomarkers to force
  AUC above 0.5
- **Matrix-native computation** — screen thousands of biomarkers in one call
- **DeLong and bootstrap inference** with multiplicity adjustment
- **Paired comparisons** on common subjects with safety limits
- **Bootstrap rank stability** — quantify reproducibility of top biomarkers
- **Two simulation engines** — binormal (sequential) and latent probit
  (joint multivariate normal)
- **Publication-quality visualizations** via `ggplot2`

---

## 1. Generating Simulated Biomarker Data

### 1.1 Latent Probit Model (Recommended)

The `generate_data_probit()` function models biomarkers and outcome through a
**joint multivariate normal** distribution with a thresholded latent variable.
This simultaneously controls **both** the between-biomarker correlations
**and** each biomarker's AUC against the outcome.

```{r sim_probit}
set.seed(42)

# Four biomarkers with decreasing AUCs, moderate correlations
sim <- generate_data_probit(
  n = 500,
  target_aucs = c(0.90, 0.80, 0.70, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.4, 0.2, 0.1,
    0.4, 1.0, 0.3, 0.2,
    0.2, 0.3, 1.0, 0.3,
    0.1, 0.2, 0.3, 1.0
  ), 4, 4),
  prevalence = 0.3
)

# Compare targets to achieved values
data.frame(
  Target   = sim$target_aucs,
  Achieved = round(sim$achieved_aucs, 4)
)

# Achieved correlation matrix
round(sim$achieved_correlations, 3)
```

### 1.2 Binormal Sequential Model

`generate_data_analytical()` uses the equal-variance binormal model,
building biomarkers sequentially from highest to lowest AUC.
Here AUC determines the mean separation; correlation is inherited
through sequential decomposition.

```{r sim_binormal}
sim2 <- generate_data_analytical(
  n = 500,
  prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.3, 0.1,
    0.3, 1.0, 0.2,
    0.1, 0.2, 1.0
  ), 3, 3)
)

round(sim2$achieved_aucs, 4)
```

### 1.3 Single AUC Vector

`generate_auc_vector()` constructs a score vector whose **exact empirical AUC**
matches a target value (up to the discrete grid of attainable AUCs).

```{r sim_auc_vec}
y <- rbinom(200, 1, 0.3)
out <- generate_auc_vector(y, target_auc = 0.80)
out$achieved_auc
```

---

## 2. Matrix Screening

The primary function `aucmat()` screens every biomarker in a numeric matrix
against a binary outcome, reporting direction-preserving AUCs, confidence
intervals, and multiplicity-adjusted p-values.

```{r screen}
# Extract predictor matrix and outcome
X <- as.matrix(sim$data[, 1:4])
y <- sim$data$truth

# Screen with DeLong CIs and BH adjustment
fit <- aucmat(X, y, ci = "delong", adjust = "BH")

# Compact display
print(fit)
```

The `summary()` method provides a detailed overview:

```{r summary}
summary(fit)
```

### Result columns

| Column | Description |
|--------|-------------|
| `auc_raw` | AUC on the observed direction (may be < 0.5) |
| `auc_strength` | $0.5 + |\text{auc\_raw} - 0.5|$ — discrimination strength |
| `effect_direction` | `higher_in_positive` or `lower_in_positive` |
| `std_error` | DeLong or bootstrap standard error |
| `conf_low`, `conf_high` | Confidence interval for `auc_raw` |
| `p_value`, `q_value` | Raw and multiplicity-adjusted p-values |
| `rank` | Rank by `auc_strength` (descending) |
| `status` | `ok`, `constant`, `insufficient_positive`, etc. |

---

## 3. Visualization

### 3.1 Rank Plot

Ordered discrimination strengths with optional CI error bars:

```{r plot_rank, fig.width=7, fig.height=4}
plot_auc_rank(fit, n_label = 4)
```

### 3.2 Volcano Plot

Effect magnitude versus statistical evidence:

```{r plot_volcano, fig.width=7, fig.height=4}
plot_auc_volcano(fit, q_cutoff = 0.05)
```

### 3.3 Forest Plot

Selected AUCs with confidence intervals:

```{r plot_forest, fig.width=7, fig.height=3}
plot_auc_forest(fit, n = 4)
```

### 3.4 ROC Curves

ROC curves for selected top biomarkers:

```{r plot_roc, fig.width=5, fig.height=5}
plot_roc_top(fit, X = X, y = y, biomarkers = head(fit$results$biomarker, 3))
```

---

## 4. Paired Comparisons

`compare_auc()` compares biomarkers on the **same subjects** using DeLong's
test for paired AUCs.  It enforces a safety limit on the number of pairs
to prevent accidental quadratic computation.

```{r compare}
# Compare top 3 biomarkers (3 choose 2 = 3 pairs)
cmp <- compare_auc(fit, X, y, top_n = 3)
print(cmp)
```

```{r compare_ref}
# Or compare all others to a reference biomarker
cmp_ref <- compare_auc(fit, X, y, reference = "X1")
print(cmp_ref)
```

---

## 5. Rank Stability

`auc_stability()` uses stratified bootstrap resampling to assess how
stable the rankings are under sampling variability.

```{r stability, fig.width=8, fig.height=4}
set.seed(123)
stab <- auc_stability(X, y, times = 200, top_k = c(3, 5), seed = 42)

# Top-k probabilities
head(stab$top_k_probs, 10)

# Stability plot
plot_auc_stability(stab, n_label = 8)
```

The plot shows median bootstrap ranks with interquartile ranges.
Narrow intervals indicate stable rankings; wide intervals suggest
the biomarker's rank depends heavily on the specific sample.

---

## 6. Missing Data

`aucmat()` handles missing biomarker values via two strategies:

- **`featurewise`** (default): each biomarker uses all subjects with
  observed values for that biomarker — maximises available data
- **`complete`**: uses only subjects with complete observations across
  **all** biomarkers — ensures directly comparable sample sizes

```{r missing}
# Introduce missing values
X_na <- X
X_na[sample(length(X_na), 50)] <- NA

fit_na <- aucmat(X_na, y, ci = "none", na_action = "featurewise")
head(fit_na$results[, c("biomarker", "n_used", "n_missing", "missing_fraction")])
```

Outcome-dependent imputation is not performed internally — impute your
data before calling `aucmat()`.

---

## 7. Reproducibility

All functions accept a `seed` argument for reproducibility:

```{r repro}
fit1 <- aucmat(X, y, ci = "bootstrap", boot_n = 500, seed = 123)
fit2 <- aucmat(X, y, ci = "bootstrap", boot_n = 500, seed = 123)
identical(fit1$results$auc_raw, fit2$results$auc_raw)
```

---

## Getting Help

```r
?aucmat
?generate_data_probit
?compare_auc
?auc_stability
?plot_auc_rank
```
