---
title: "Practical Biomarker Screening with aucmat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{Practical Biomarker Screening with aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, message = FALSE, warning = FALSE)
```

```{r setup}
library(aucmat)
```

## Overview

This vignette walks through a realistic biomarker screening analysis using
`aucmat`, from simulation through to reporting. We follow a typical omics
workflow: generate data with known properties, screen biomarkers, compare
leading candidates, validate via cross-validation, combine into a panel,
and assess power for a follow-up study.

---

## 1. Simulate Realistic Biomarker Data

We simulate 500 subjects (30% cases) with 20 biomarkers. The first 5 have
moderate-to-strong signals (AUC 0.65–0.85); the remaining 15 are noise
(AUC 0.50–0.55). Biomarkers within functional groups are correlated.

```{r sim_data}
set.seed(42)
n_biomarkers <- 20
signal_aucs <- c(0.85, 0.78, 0.72, 0.68, 0.65)
noise_aucs <- runif(15, 0.50, 0.55)
target_aucs <- c(signal_aucs, noise_aucs)

sim <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = target_aucs,
  correlation = 0.15, structure = "exchangeable",
  seed = 42
)

X <- as.matrix(sim$data[, 1:20])
y <- sim$data$truth
```

---

## 2. Screen All Biomarkers

```{r screen}
fit <- aucmat(X, y, ci = "delong", adjust = "BH")
summary(fit)
```

### Top biomarkers

```{r top}
head(as.data.frame(fit), 8)
```

---

## 3. Hurdle-AUC for Zero-Inflated Biomarkers

Now let's add zero-inflation to simulate scRNA-seq-like data and compare
standard AUC against the two-stage Hurdle-AUC.

```{r hurdle_data}
set.seed(123)
sim_hur <- simulate_hurdle_auc(
  n = 400, prevalence = 0.3,
  target_hurdle_aucs = c(0.85, 0.72, 0.60, 0.55, 0.52),
  zero_rate_neg = c(0.55, 0.30, 0.70, 0.80, 0.90),
  zero_rate_pos = c(0.25, 0.10, 0.60, 0.75, 0.88)
)
Xh <- as.matrix(sim_hur$data[, 1:5])
yh <- sim_hur$data$truth

# Standard AUC
fit_std <- aucmat(Xh, yh, ci = "none")

# Hurdle AUC
fit_hur <- hurdle_auc(Xh, yh)
```

```{r hurdle_compare}
# Side-by-side comparison
comp <- data.frame(
  Biomarker     = paste0("X", 1:5),
  Standard_AUC  = round(fit_std$results$auc_strength, 3),
  Hurdle_AUC    = round(fit_hur$results$hurdle_auc, 3),
  Zero_AUC      = round(fit_hur$results$zero_auc, 3),
  Nonzero_AUC   = round(fit_hur$results$nonzero_auc, 3),
  Zero_Rate     = round(fit_hur$results$zero_rate_total, 3)
)
comp$Improvement <- round((comp$Hurdle_AUC - comp$Standard_AUC) / comp$Standard_AUC * 100, 1)
comp
```

```{r hurdle_diag, fig.width=7, fig.height=4.5}
plot_hurdle_diagnostics(fit_hur)
```

The Hurdle-AUC decomposes discrimination into zero-inflation signal and
magnitude signal, recovering AUCs that standard screening misses on
zero-heavy biomarkers.

---

## 4. Visualize Results

### Rank plot

```{r rank_plot, fig.width=8, fig.height=4}
plot_auc_rank(fit, n_label = 10)
```

### Volcano plot

```{r volcano_plot, fig.width=8, fig.height=4}
plot_auc_volcano(fit, q_cutoff = 0.05)
```

### Forest plot of top biomarkers

```{r forest_plot, fig.width=7, fig.height=3}
plot_auc_forest(fit, n = 6)
```

---

## 5. Compare Leading Candidates

```{r compare}
# All pairs among the top 5
cmp <- compare_auc(fit, X, y, top_n = 5, adjust = "BH")
print(cmp)
```

### Forest plot of pairwise differences

```{r compare_plot, fig.width=8, fig.height=5}
plot(cmp)
```

### Global test

```{r global}
compare_auc_global(fit, X, y, biomarkers = head(fit$results$biomarker, 5))
```

---

## 6. Assess Rank Stability

```{r stability}
stab <- auc_stability(X, y, times = 500, seed = 123)
print(stab)
```

```{r stability_plot, fig.width=8, fig.height=5}
plot_auc_stability(stab, n_label = 12)
```

---

## 7. Cross-Validated Screening

Same-data AUCs are optimistically biased. Cross-validation gives honest
out-of-fold estimates:

```{r cv}
cv <- cv_aucmat(X, y, n_folds = 5, seed = 123)
print(cv)

# Compare: in-sample vs CV
head(data.frame(
  biomarker = cv$results$biomarker,
  auc_in_sample = round(cv$results$auc_in_sample, 3),
  auc_cv = round(cv$results$auc_cv, 3),
  optimism = round(cv$results$optimism, 3)
), 8)
```

---

## 8. Build a Multivariable Panel

Combine the top biomarkers into a single score. `fit_auc_panel()` reports
both an in-sample AUC and an **honest, nested** cross-validated AUC:
penalty selection and centering/scaling are re-derived from each fold's
training portion alone, so the held-out fold never leaks into the model
evaluated on it.

```{r panel}
# Select top 4 biomarkers by CV AUC
top4 <- head(cv$results$biomarker, 4)
X_panel <- X[, top4, drop = FALSE]

panel <- fit_auc_panel(X_panel, y, method = "ridge", n_folds = 5, seed = 123)
print(panel)
```

```{r panel_plot, fig.width=7, fig.height=3}
plot(panel)
```

### 8.1 Other combination methods

Ridge is one of six options. `"su_liu"` is a closed-form linear
combination that maximizes AUC under the multivariate-normal (binormal)
model -- fast, no `glmnet` dependency, and a natural fit alongside this
package's own binormal simulation engine. `"unweighted"` sums
direction-aligned, standardized biomarkers with equal weight and fits
nothing at all -- a useful baseline for judging whether the fitted methods
are earning their complexity.

```{r panel_methods}
panel_su <- fit_auc_panel(X_panel, y, method = "su_liu", n_folds = 5, seed = 123)
panel_uw <- fit_auc_panel(X_panel, y, method = "unweighted", n_folds = 5, seed = 123)

data.frame(
  method   = c("ridge", "su_liu", "unweighted"),
  auc_train = c(panel$auc_train, panel_su$auc_train, panel_uw$auc_train),
  auc_cv    = c(panel$auc_cv, panel_su$auc_cv, panel_uw$auc_cv)
)
```

### 8.2 Panel ROC vs. individual biomarkers

```{r panel_roc, fig.width=7, fig.height=5}
plot_roc_panel(panel, X_panel, y)
```

### 8.3 Does combining actually help?

`compare_panel_auc()` tests the panel's honest cross-validated score
against each individual biomarker on the same subjects -- a confidence
interval and p-value for whether combining helped, rather than two AUC
numbers to eyeball.

```{r panel_compare}
compare_panel_auc(panel, X_panel, y)
```

---

## 9. Plan a Follow-Up Study

What sample size is needed to detect AUC = 0.70 with 80% power?

```{r power}
pwr <- power_auc_matrix(
  target_aucs = 0.70,
  power = 0.80,
  prevalence = 0.3,
  alpha = 0.05
)
print(pwr)
```

---

## Summary

| Step | Function | Key output |
|------|----------|------------|
| Simulate | `simulate_auc_matrix()` | Biomarker matrix + outcome |
| Screen | `aucmat()` | Direction-preserving AUCs with q-values |
| Visualize | `plot_auc_rank()`, `plot_auc_volcano()`, `plot_auc_forest()` | Publication-ready plots |
| Compare | `compare_auc()` | Paired differences with CIs |
| Global test | `compare_auc_global()` | Omnibus Wald test |
| Stability | `auc_stability()` | Bootstrap rank distributions |
| Cross-validate | `cv_aucmat()` | Out-of-fold AUCs, optimism |
| Panel | `fit_auc_panel()` | Combined biomarker score, honest nested CV AUC |
| Panel ROC | `plot_roc_panel()` | Panel vs. individual-biomarker ROC curves |
| Panel comparison | `compare_panel_auc()` | Does the panel beat its components? |
| Power | `power_auc_matrix()` | Sample size for next study |
