---
title: "Practical Biomarker Screening with aucmat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{Practical Biomarker Screening with aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, message = FALSE, warning = FALSE)
```

```{r setup}
library(aucmat)
```

## Overview

This vignette walks through a realistic biomarker screening analysis using
`aucmat`, from simulation through to reporting. We follow a typical omics
workflow: generate data with known properties, screen biomarkers, compare
leading candidates, validate via cross-validation, combine into a panel,
and assess power for a follow-up study.

---

## 1. Simulate Realistic Biomarker Data

We simulate 500 subjects (30% cases) with 20 biomarkers. The first 5 have
moderate-to-strong signals (AUC 0.65–0.85); the remaining 15 are noise
(AUC 0.50–0.55). Biomarkers within functional groups are correlated.

```{r sim_data}
set.seed(42)
n_biomarkers <- 20
signal_aucs <- c(0.85, 0.78, 0.72, 0.68, 0.65)
noise_aucs <- runif(15, 0.50, 0.55)
target_aucs <- c(signal_aucs, noise_aucs)

sim <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = target_aucs,
  correlation = 0.15, structure = "exchangeable",
  seed = 42
)

X <- as.matrix(sim$data[, 1:20])
y <- sim$data$truth
```

---

## 2. Screen All Biomarkers

```{r screen}
fit <- aucmat(X, y, ci = "delong", adjust = "BH")
summary(fit)
```

### Top biomarkers

```{r top}
head(as.data.frame(fit), 8)
```

---

## 3. Visualize Results

### Rank plot

```{r rank_plot, fig.width=8, fig.height=4}
plot_auc_rank(fit, n_label = 10)
```

### Volcano plot

```{r volcano_plot, fig.width=8, fig.height=4}
plot_auc_volcano(fit, q_cutoff = 0.05)
```

### Forest plot of top biomarkers

```{r forest_plot, fig.width=7, fig.height=3}
plot_auc_forest(fit, n = 6)
```

---

## 4. Compare Leading Candidates

```{r compare}
# All pairs among the top 5
cmp <- compare_auc(fit, X, y, top_n = 5, adjust = "BH")
print(cmp)
```

### Forest plot of pairwise differences

```{r compare_plot, fig.width=8, fig.height=5}
plot(cmp)
```

### Global test

```{r global}
compare_auc_global(fit, X, y, biomarkers = head(fit$results$biomarker, 5))
```

---

## 5. Assess Rank Stability

```{r stability}
stab <- auc_stability(X, y, times = 500, seed = 123)
print(stab)
```

```{r stability_plot, fig.width=8, fig.height=5}
plot_auc_stability(stab, n_label = 12)
```

---

## 6. Cross-Validated Screening

Same-data AUCs are optimistically biased. Cross-validation gives honest
out-of-fold estimates:

```{r cv}
cv <- cv_aucmat(X, y, n_folds = 5, seed = 123)
print(cv)

# Compare: in-sample vs CV
head(data.frame(
  biomarker = cv$results$biomarker,
  auc_in_sample = round(cv$results$auc_in_sample, 3),
  auc_cv = round(cv$results$auc_cv, 3),
  optimism = round(cv$results$optimism, 3)
), 8)
```

---

## 7. Build a Multivariable Panel

Combine the top biomarkers into a single score:

```{r panel}
# Select top 4 biomarkers by CV AUC
top4 <- head(cv$results$biomarker, 4)
X_panel <- X[, top4, drop = FALSE]

panel <- fit_auc_panel(X_panel, y, method = "ridge", n_folds = 5, seed = 123)
print(panel)
```

```{r panel_plot, fig.width=7, fig.height=3}
plot(panel)
```

---

## 8. Plan a Follow-Up Study

What sample size is needed to detect AUC = 0.70 with 80% power?

```{r power}
pwr <- power_auc_matrix(
  target_aucs = 0.70,
  power = 0.80,
  prevalence = 0.3,
  alpha = 0.05
)
print(pwr)
```

---

## Summary

| Step | Function | Key output |
|------|----------|------------|
| Simulate | `simulate_auc_matrix()` | Biomarker matrix + outcome |
| Screen | `aucmat()` | Direction-preserving AUCs with q-values |
| Visualize | `plot_auc_rank()`, `plot_auc_volcano()`, `plot_auc_forest()` | Publication-ready plots |
| Compare | `compare_auc()` | Paired differences with CIs |
| Global test | `compare_auc_global()` | Omnibus Wald test |
| Stability | `auc_stability()` | Bootstrap rank distributions |
| Cross-validate | `cv_aucmat()` | Out-of-fold AUCs, optimism |
| Panel | `fit_auc_panel()` | Combined biomarker score |
| Power | `power_auc_matrix()` | Sample size for next study |
