---
title: "Real-Data Application: BeatMG Proteomics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{Real-Data Application: BeatMG Proteomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
have_dt <- requireNamespace("DT", quietly = TRUE)
```

```{r setup}
library(aucmat)
```

## Overview

Every other vignette in this package uses simulated data with known ground
truth -- useful for validating methodology, but it can't show what `aucmat()`
actually finds in a real biomarker matrix. This vignette does that, using
plasma Olink proteomics from **BeatMG** (NeuroNEXT NN102), a randomized,
double-blind, placebo-controlled trial of rituximab for myasthenia gravis.

Rituximab is an anti-CD20 monoclonal antibody that depletes B cells. That
gives us a real, mechanistically falsifiable question to screen for: *does
the plasma proteome show rituximab's effect, and if so, through which
proteins?* Two datasets ship with `aucmat`:

- `beatmg_baseline`: 48 subjects, pre-treatment (Timepoint 1)
- `beatmg_ontreatment`: 44 subjects, on-treatment (Timepoint 3 -- the latest
  collection with enough samples for AUC inference; Timepoint 4 has only 5)

Both are Olink Inflammation + Inflammation_II panels (730 proteins),
quality-controlled (`?beatmg_baseline` for exact QC criteria). Randomization
means baseline should show *no* systematic biomarker--arm association --
that makes it a real-data negative control, not just a second discovery run.

---

## 1. The Raw Data

```{r load_data}
data(beatmg_baseline)
data(beatmg_ontreatment)
dim(beatmg_baseline)
dim(beatmg_ontreatment)
table(beatmg_baseline$arm)
```

```{r raw_table_dt, eval=have_dt}
DT::datatable(
  beatmg_ontreatment[, 1:12],
  caption = "On-treatment data (first 10 proteins shown; 730 total columns)",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

```{r raw_table_fallback, eval=!have_dt}
head(beatmg_ontreatment[, 1:12])
```

---

## 2. Per-Biomarker Descriptive Statistics

Before screening, look at what's actually in the matrix: distribution of
each protein's NPX values, plus its raw AUC against treatment arm.

```{r biomarker_stats}
protein_cols <- setdiff(names(beatmg_ontreatment), c("SampleID", "arm"))
X <- as.matrix(beatmg_ontreatment[, protein_cols])
y <- beatmg_ontreatment$arm

fit <- aucmat(X, y, positive = "Rituximab", ci = "delong", adjust = "BH")

desc <- data.frame(
  biomarker = protein_cols,
  mean      = colMeans(X),
  sd        = apply(X, 2, stats::sd),
  min       = apply(X, 2, min),
  max       = apply(X, 2, max)
)
biomarker_summary <- merge(
  desc,
  fit$results[, c("biomarker", "auc_raw", "conf_low", "conf_high",
                   "p_value", "q_value")],
  by = "biomarker"
)
biomarker_summary <- biomarker_summary[order(-abs(biomarker_summary$auc_raw - 0.5)), ]
biomarker_summary[, -1] <- round(biomarker_summary[, -1], 4)
```

```{r biomarker_stats_dt, eval=have_dt}
DT::datatable(
  biomarker_summary,
  caption = "Per-biomarker NPX distribution and AUC (ranked by |AUC - 0.5|)",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

```{r biomarker_stats_fallback, eval=!have_dt}
head(biomarker_summary, 10)
```

---

## 3. Screening: Baseline vs. On-Treatment

### 3.1 Baseline (randomization check)

```{r baseline_screen}
fit_baseline <- aucmat(
  as.matrix(beatmg_baseline[, protein_cols]), beatmg_baseline$arm,
  positive = "Rituximab", ci = "delong", adjust = "BH"
)
sum(fit_baseline$results$q_value < 0.05, na.rm = TRUE)
```

Zero proteins reach `q < 0.05` out of 730 tested. That's the expected result
under randomization -- and it rules out an easy alternative explanation for
whatever the on-treatment screen finds (a systematic batch/plate artifact
would show up at every timepoint, not appear only on-treatment).

### 3.2 On-treatment (discovery)

```{r ontreatment_screen}
print(fit)
sum(fit$results$q_value < 0.05, na.rm = TRUE)
hits <- fit$results[!is.na(fit$results$q_value) & fit$results$q_value < 0.05, ]
hits[order(hits$auc_raw), c("biomarker", "auc_raw", "conf_low", "conf_high", "q_value")]
```

Eight proteins are significant at on-treatment, all with `auc_raw < 0.5` --
systematically *lower* under rituximab. Six of the eight (`FCRL2`, `CD22`,
`TNFRSF13C`/BAFF-R, `TNFRSF13B`/TACI, `CD79B`, `CD72`) are canonical B-cell
surface or B-cell-survival-signaling proteins. Finding exactly these
suppressed in a blinded, randomized comparison is the textbook
pharmacodynamic signature of anti-CD20 B-cell depletion -- not an
after-the-fact story fit to noise.

Cross-checking against bootstrap CI (small per-arm n here, ~21-23, is below
where DeLong's asymptotic coverage is fully reliable):

```{r bootstrap_cross_check}
fit_boot <- aucmat(X, y, positive = "Rituximab", ci = "bootstrap",
                    boot_n = 1000, adjust = "BH", seed = 42)
sum(fit_boot$results$q_value < 0.05, na.rm = TRUE)
```

Same eight proteins, in close numerical agreement -- DeLong is not
misleading us here despite the small sample.

---

## 4. Validation: Is the Ranking Stable?

A single screen is not evidence of a stable ranking. `auc_stability()`
resamples (stratified by arm) and reports how often each protein actually
lands in the top-k.

```{r stability, fig.width=8, fig.height=5}
stab <- auc_stability(X, y, positive = "Rituximab",
                       times = 1000, top_k = c(8, 15, 25), seed = 42)
print(stab)
plot_auc_stability(stab, n_label = 10)
```

The eight significant hits split into two stability tiers: `FCRL2`, `CD22`,
`IL17F`, `TNFRSF13C` are rock-stable (top-8 in essentially every resample);
`TNFRSF13B` and `SPINK2` are moderately stable; `CD79B` and `CD72` are
fragile -- most resamples do *not* reproduce them in the top 8. Report the
last two with more caution than the first six.

---

## 5. Visualization

```{r rank_plot}
plot_auc_rank(fit, n_label = 10)
```

```{r volcano_plot}
plot_auc_volcano(fit, n_label = 10, q_cutoff = 0.05)
```

---

## Scope and Limitations

This analysis shows rituximab's expected pharmacodynamic effect on
circulating B-cell markers -- it is **not** a clinical efficacy or
treatment-response result. `beatmg_baseline`/`beatmg_ontreatment` carry only
randomized arm, not the trial's clinical endpoints (MG-ADL, QMG,
prednisone-sparing outcome), which are not part of this data release. The
datasets are also an available-case subset (44-48 of 52 randomized
participants, two of four Olink panels) rather than the full cohort or
proteome.

## Summary

| Step | Function | Key output |
|------|----------|------------|
| Load real data | `data(beatmg_baseline)`, `data(beatmg_ontreatment)` | Subject x protein NPX matrix + randomized arm |
| Screen | `aucmat()` | Direction-preserving AUCs with DeLong/bootstrap CIs and q-values |
| Validate | `auc_stability()` | Bootstrap rank-stability, top-k selection frequency |
| Visualize | `plot_auc_rank()`, `plot_auc_volcano()`, `plot_auc_stability()` | Publication-ready plots |
