---
title: "Simulating Biomarker Data in aucmat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{Simulating Biomarker Data in aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r setup}
library(aucmat)
```

## Overview

`aucmat` provides seven simulation functions for generating biomarker data
with controlled properties. This vignette explains the mathematical
foundations, strengths, and limitations of each.

### Quick Reference

| Function | AUC | Correlation | Approach | Best for |
|----------|-----|-------------|----------|----------|
| `simulate_auc_matrix()` | Target via $\delta$ | Named structures | Class-conditional MVN | General use |
| `generate_data_probit()` | Target via $\rho$ | Full matrix | Latent MVN | Free AUC + correlation |
| `generate_data_analytical()` | Target via $\delta$ | Sequential approx | Binormal decomposition | Fast validation |
| `generate_auc_vector()` | **Exact** empirical | None (post-hoc Box-Cox) | Rank construction | Exact AUC guarantees |
| `generate_auc_cor_vector()` | **Exact** empirical | **Approx** Pearson r | Rank + Box-Cox tuning | Exact AUC with correlation |
| `simulate_auc_correlation()` | MC distribution | MC distribution | Binormal draws | Understanding sampling variation |
| `validate_simulation()` | Repeat + diagnostics | Repeat + diagnostics | Independent replication | Calibration confirmation |

---

## 1. `simulate_auc_matrix()` — Recommended General-Purpose Simulator

### Mathematical Foundation

Rather than deriving the outcome from a thresholded latent variable,
`simulate_auc_matrix()` fixes the outcome $Y$ first and draws biomarkers
**conditional on class**:

$$X \mid Y = 0 \sim \mathcal{N}_p(\mu_0, \mathbf{R}), \qquad
  X \mid Y = 1 \sim \mathcal{N}_p(\mu_1, \mathbf{R})$$

Both classes share the same correlation matrix $\mathbf{R}$; only the mean
vectors differ. The per-biomarker mean shift $\delta_k = \mu_{1k} - \mu_{0k}$
uses closed-form binormal calibration:

$$\delta_k = \sqrt{2} \cdot \Phi^{-1}(\text{AUC}_k)$$

The class means are centred so the $n$-weighted grand mean is zero:

$$\mu_0 = \delta \cdot \left(-\frac{n_1}{n}\right), \qquad
  \mu_1 = \delta \cdot \left(\frac{n_0}{n}\right)$$

Because $Y$ is fixed *before* any biomarker is drawn, supplied outcomes keep
their exact values and row order, and generated outcomes get exact class
counts (default) or Bernoulli counts.

### Outcome Modes

```{r matrix_outcome_modes}
set.seed(42)

# 1. Generate outcome from n + prevalence (default: exact class counts)
sim_gen <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.3, structure = "exchangeable"
)
table(sim_gen$data$truth)  # exact 150 / 350

# 2. Bernoulli outcome (random class counts)
sim_bern <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.3, structure = "exchangeable",
  outcome_mode = "bernoulli", seed = 45
)
table(sim_bern$data$truth)  # varies around 30%

# 3. Supplied outcome (exact values preserved)
y_supplied <- rbinom(300, 1, 0.3)
sim_y <- simulate_auc_matrix(
  y = y_supplied,
  target_aucs = c(0.85, 0.75),
  correlation = 0.3, structure = "exchangeable", seed = 1
)
all(sim_y$data$truth == y_supplied)
```

### Correlation Structures

Instead of hand-building a full matrix, `structure` names a pattern:

| `structure` | Parameters | Formula for $R_{ij}$ |
|-------------|-----------|----------------------|
| `"user"` | Full matrix `correlation` | As supplied |
| `"exchangeable"` | `correlation` = $\rho$ | $\rho$ for all $i \neq j$ |
| `"ar1"` | `correlation` = $\rho$ | $\rho^{|i-j|}$ |
| `"block"` | `block_sizes`, $\rho_w$, $\rho_b$ | $\rho_w$ in block, $\rho_b$ between |

```{r matrix_structures}
# Exchangeable
sim_ex <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.4, structure = "exchangeable", seed = 1
)
cat("Exchangeable (rho = 0.4):\n")
round(sim_ex$achieved_correlation, 3)

# AR(1) — correlation decays with distance
sim_ar1 <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  correlation = 0.6, structure = "ar1", seed = 1
)
cat("\nAR(1) (rho = 0.6):\n")
round(sim_ar1$achieved_correlation, 3)

# Block — two groups of two with different within/between correlations
sim_blk <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.8, 0.7, 0.65),
  structure = "block", block_sizes = c(2, 2),
  rho_within = 0.6, rho_between = 0.1, seed = 1
)
cat("\nBlock (rho_w = 0.6, rho_b = 0.1):\n")
round(sim_blk$achieved_correlation, 3)
```

### Feasibility Diagnostics

Not every correlation matrix is positive definite. `feasibility` controls
the response:

- `"error"` (default): stops with a structured condition and quantitative
  diagnostic (eigenvalue, matrix dimensions).
- `"nearest"`: projects to the nearest positive-semidefinite matrix via
  eigenvalue clipping, reports Frobenius adjustment and maximum element-wise
  change.

```{r matrix_feasibility}
# This is infeasible: exchangeable(0.4) with p=4 is PD; exchangeable(-0.5) is not
# because the min eigenvalue of an exchangeable matrix with rho is
# (1 - rho) for p-1 multiplicities, needs rho > -1/(p-1) = -1/3 for p=4
sim_adj <- simulate_auc_matrix(
  n = 200, prevalence = 0.3,
  target_aucs = c(0.8, 0.7, 0.6, 0.55),
  correlation = -0.5, structure = "exchangeable",
  feasibility = "nearest", seed = 1
)
sim_adj$feasibility
```

### A Note on Achieved Correlation

`$achieved_correlation` is computed on **within-class centred** residuals —
each class centred on its own empirical mean before correlating. This is
what `$requested_correlation` ($\mathbf{R}$) actually parametrizes. The raw
pooled correlation of the returned columns (ignoring `truth`) is **not** a
consistent estimate of $\mathbf{R}$ whenever any AUC differs from 0.5:
mixing subpopulations with different means inflates the pooled correlation
(a mixture / Simpson's-paradox effect).

```{r matrix_correlation_note}
sim_demo <- simulate_auc_matrix(
  n = 1000, prevalence = 0.3,
  target_aucs = c(0.85, 0.75), correlation = 0.3,
  structure = "exchangeable", seed = 1
)

# Naive pooled correlation (wrong estimate of R):
cat("Naive pooled cor(X1, X2):             ",
    round(cor(sim_demo$data$X1, sim_demo$data$X2), 4), "\n")
# Correct within-class correlation (estimate of R):
cat("Within-class centred cor (correct):    ",
    round(sim_demo$achieved_correlation[1, 2], 4), "\n")
cat("Requested correlation (R):             ",
    sim_demo$requested_correlation[1, 2], "\n")
```

### Advantages

- **Fast** — closed-form mean shift, no numerical calibration loop
- **Named correlation structures** — exchangeable, AR(1), block, or
  hand-supplied matrix
- **Feasibility diagnostics** — structured errors or automatic nearest-PD
  repair with adjustment magnitude reported
- **Outcome-preserving** — supplied `y` is never altered; generated `y` gets
  exact class counts by default

### Limitations

- Equal-variance binormal assumption (both classes share the same $\mathbf{R}$)
- Correlation structure limited to common parametric patterns
- Non-normal marginals, batch effects, outliers, and missingness not yet
  supported (planned for later milestones)

---

## 2. `validate_simulation()` — Calibration Check

A single draw is not evidence a simulator is calibrated. This function
repeats a `simulate_auc_matrix()` specification across many independent
replicates with deterministic seed streams (seed + replicate_index − 1) and
reports:

| Metric | AUC | Correlation |
|--------|-----|-------------|
| Bias | `colMeans(achieved) - target` | Per upper-triangle pair |
| RMSE | `sqrt(colMeans((achieved - target)^2))` | Per pair |
| MC SE | `sd / sqrt(n_ok)` — estimation uncertainty | Per pair |
| Hit rate | Fraction within tolerance of target | Per pair |

```{r validate_example}
val <- validate_simulation(
  n = 300, prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  correlation = 0.3, structure = "exchangeable",
  times = 100, seed = 1
)
print(val)
```

### Interpreting the Output

- **Bias** should be small relative to `auc_mc_se` (Monte Carlo standard
  error). If bias is consistently much larger than MC SE across replicates,
  the simulator has a systematic calibration error.
- **RMSE** captures both bias and variance. At small `times`, RMSE is
  dominated by sampling variation; at large `times`, remaining RMSE reflects
  bias.
- **Hit rate** depends on `tolerance`. With `tolerance$auc = 0.02`, roughly
  95% of samples should hit when the Monte Carlo SE of the achieved AUC is
  about 0.01 (i.e., `n` around 300-500). Hit rates below ~30% with
  `times = 100` can indicate a calibration problem or simply that `tolerance`
  is too tight for the chosen `n`.

### Failure Handling

`n_failed` records replicates where `simulate_auc_matrix()` threw an error.
Requires at least 2 successful replicates. Successful and failed counts are
always reported.

---

## 3. Latent Probit Model (`generate_data_probit`)

### Mathematical Foundation

The binary outcome $Y$ arises from a latent continuous variable $Y^*$:

$$Y^* \sim \mathcal{N}(0, 1), \qquad
  Y = \mathbf{1}\{Y^* > \tau\}, \qquad
  \tau = \Phi^{-1}(1 - \pi)$$

Biomarkers $(X_1, \ldots, X_p, Y^*)$ are jointly multivariate normal:

$$\begin{pmatrix} X_1 \\ \vdots \\ X_p \\ Y^* \end{pmatrix} \sim
  \mathcal{N}_{p+1}\left(
  \mathbf{0},
  \begin{pmatrix}
  \mathbf{R} & \boldsymbol{\rho} \\
  \boldsymbol{\rho}^\top & 1
  \end{pmatrix}
  \right)$$

where $\mathbf{R}$ is the $p \times p$ between-biomarker correlation
matrix and $\rho_k = \text{Cor}(X_k, Y^*)$ governs the biomarker-outcome
relationship.

### AUC–$\rho$ Relationship

For prevalence $\pi$, the AUC is a strictly increasing function of $\rho_k$:

$$\text{AUC}(\rho_k, \pi) = f(\rho_k, \pi)$$

$f$ involves the bivariate normal CDF (no closed form). Calibration uses
bisection over a large sample ($n = 100,000$ by default).

### Example

```{r probit_example}
set.seed(42)
sim_p <- generate_data_probit(
  n = 500,
  target_aucs = c(0.90, 0.80, 0.70, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.4, 0.2, 0.1,
    0.4, 1.0, 0.3, 0.2,
    0.2, 0.3, 1.0, 0.3,
    0.1, 0.2, 0.3, 1.0
  ), 4, 4),
  prevalence = 0.3
)

# Target vs achieved AUCs
data.frame(
  Target   = sim_p$target_aucs,
  Achieved = round(sim_p$achieved_aucs, 4),
  Latent_rho = round(sim_p$latent_rhos, 4)
)
```

### Advantages

- **Both AUC and correlation are free parameters** — independent control
- **Natural framework** — single $(p+1) \times (p+1)$ matrix captures all
  relationships
- **Positive definiteness** checked and repaired automatically

### Limitations

- Numerical calibration is expensive ($n_{\text{cal}} = 100,000$ draws per
  AUC, bisection to $10^{-6}$ tolerance)
- Not every $(p+1) \times (p+1)$ correlation matrix is positive definite
- Requires `mvtnorm` for multivariate normal draws

---

## 4. Sequential Binormal Model (`generate_data_analytical`)

### Mathematical Foundation

Biomarkers are constructed from **highest to lowest AUC**:

1. The first biomarker $X_1$ is drawn with the full mean separation
   $\delta_1 = \sqrt{2} \cdot \Phi^{-1}(\text{AUC}_1)$.

2. For $k \geq 2$, $X_k$ is built in two parts:
   - **Correlated component**: linear combination of $X_1, \ldots, X_{k-1}$
     weighted by $\mathbf{R}_{k, 1:k-1} \cdot \mathbf{R}_{1:k-1, 1:k-1}^{-1}$
   - **Residual component**: independent normal carrying the remaining mean
     separation needed to hit $\text{AUC}_k$ after accounting for what the
     correlated component already contributes

3. $X_k$ is centred and scaled to unit variance.

### AUC–Correlation Link

In this model, for a given prevalence $\pi$, the point-biserial correlation
between $X_k$ and $Y$ is **determined** by the AUC:

$$r_{pb} = \frac{\delta_k \cdot \sqrt{\pi(1-\pi)}}
                {\sqrt{\delta_k^2 \cdot \pi(1-\pi) + 1}}$$

| AUC | $r_{pb}$ at $\pi = 0.3$ | $r_{pb}$ at $\pi = 0.5$ |
|-----|--------------------------|--------------------------|
| 0.60 | 0.12 | 0.16 |
| 0.70 | 0.23 | 0.31 |
| 0.80 | 0.36 | 0.47 |
| 0.90 | 0.51 | 0.64 |
| 0.95 | 0.60 | 0.73 |

### Example

```{r binormal_example}
set.seed(42)
sim_b <- generate_data_analytical(
  n = 500,
  prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.3, 0.1,
    0.3, 1.0, 0.2,
    0.1, 0.2, 1.0
  ), 3, 3)
)

data.frame(
  Biomarker   = paste0("X", 1:3),
  Target_AUC  = c(0.85, 0.75, 0.65),
  Achieved_AUC = round(sim_b$achieved_aucs, 4)
)
```

### Advantages

- **Fast** — closed-form calibration, no numerical search
- **Simple** — interpretable equal-variance binormal model
- **Common in literature** — Hanley & McNeil (1982), Obuchowski (1997)

### Limitations

- **AUC and correlation are linked** — cannot independently specify both
- Equal-variance assumption may be restrictive
- Sequential construction means lower-AUC biomarkers inherit residuals from
  higher-AUC ones

---

## 5. Rank-Based Construction (`generate_auc_vector`, `generate_auc_cor_vector`)

### `generate_auc_vector()` — Exact Empirical AUC

The empirical AUC depends **only on the rank ordering** of scores. On a
finite sample with $n_1$ positives and $n_0$ negatives, achievable AUC
values form a discrete grid with step $\Delta_{\text{AUC}} = 1/(n_1 n_0)$.

```{r rank_exact}
y <- rbinom(200, 1, 0.3)
out <- generate_auc_vector(y, target_auc = 0.80)

c(Requested = out$requested_auc, Achieved = out$achieved_auc,
  AUC_step = out$auc_step)
```

When the requested AUC is not exactly attainable, `unattainable = "nearest"`
uses the closest grid value and warns.

### `generate_auc_cor_vector()` — AUC + Correlation

After constructing ranks for the target AUC, a strictly increasing Box-Cox
transform is tuned to move the Pearson correlation toward a target without
changing the AUC (since monotone transforms preserve ranks, hence preserve AUC).

```{r rank_cor}
out_cor <- generate_auc_cor_vector(
  y, target_auc = 0.80, target_cor = 0.45,
  shuffle_within_class = FALSE
)

c(AUC = out_cor$achieved_auc, Correlation = out_cor$achieved_cor,
  lambda = out_cor$lambda,
  cor_min = out_cor$cor_range[1], cor_max = out_cor$cor_range[2])
```

Not every (AUC, correlation) pair is achievable — the Box-Cox family can
only move correlation within a bounded range $[r_{\min}, r_{\max}]$.

### `simulate_auc_correlation()` — Monte Carlo Distribution

Repeatedly draws from the binormal model and records achieved (AUC, r) pairs,
showing the sampling distribution:

```{r rank_mc}
mc <- simulate_auc_correlation(
  y, target_auc = c(0.7, 0.8, 0.9), n_sim = 500, seed = 123
)

# Empirical mean AUCs (should be near targets)
aggregate(auc ~ target_auc, data = mc$results, mean)

# Mean shift for each target AUC
mc$mean_shift
```

---

## 6. Gaussian Copula + Perturbation (`simulate_auc_copula`)

### Mathematical Foundation

The copula simulator cleanly separates two orthogonal concerns:

**Phase 1 — Gaussian Copula:** For each class independently, biomarkers are
drawn from MVN(0, R) then rank-transformed to uniforms and probit-transformed
back to normal scores.  This copula step ensures the rank correlation
structure $\mathbf{R}$ is preserved exactly, regardless of the marginal
distributions.

**Phase 2 — Iterative AUC Perturbation:** For each biomarker whose achieved
AUC deviates from the target, a small class-conditional mean shift is applied.
The shift magnitude follows the binormal calibration
$\delta = \sqrt{2} \cdot \Phi^{-1}(\text{AUC})$ and decays with each
iteration (step size $\times$ `step_decay`${}^\text{iter}$).  Because only
means are shifted (not correlations), the correlation structure from Phase 1
is minimally disturbed.

### Example

```{r copula_example}
set.seed(42)
sim_cop <- simulate_auc_copula(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.4, 0.2,
    0.4, 1.0, 0.3,
    0.2, 0.3, 1.0
  ), 3, 3)
)

data.frame(
  Target   = sim_cop$target_aucs,
  Phase1   = round(sim_cop$phase1_aucs, 3),
  Achieved = round(sim_cop$achieved_aucs, 3)
)

cat("\nAchieved correlation:\n")
round(sim_cop$achieved_correlation, 3)
```

### Advantages

- **Strong correlation preservation** — copula rank-transform preserves
  the target $\mathbf{R}$ better than sequential methods
- **No AUC-correlation tradeoff** — Phase 2 perturbs only means
- **Flexible** — can be extended to non-normal marginals via ECDF
  (future milestone)

### Limitations

- Requires `mvtnorm` for multivariate normal draws
- Iterative perturbation adds computational cost vs closed-form methods
- Currently limited to normal marginals (ECDF extension deferred)

---

## 7. Hurdle Model for Zero-Inflated Data (`simulate_hurdle_auc`)

### Mathematical Foundation

In scRNA-seq and similar assays, a large fraction of measurements are exactly
zero.  The standard binormal model cannot represent this: it would require
the tail of a normal distribution below zero, producing implausible negative
values.  The hurdle model separates the data-generating process into two
stages:

**Stage 1 — Zero Hurdle:**
$$P(X_j = 0 \mid y) = \text{logistic}(\beta_0 + \beta_1 y)$$
The difference in zero rates between classes provides discriminative signal.

**Stage 2 — Magnitude (if expressed):**
$$X_j \mid X_j > 0, y \sim \text{LogNormal}(\mu_y, \sigma^2)$$
with $\mu_1 - \mu_0$ calibrated to the target non-zero AUC via the binormal
shift $\delta = \sqrt{2} \cdot \Phi^{-1}(\text{AUC}_{\text{nonzero}})$.

### Example

```{r hurdle_example}
set.seed(1)
sim_hur <- simulate_hurdle_auc(
  n = 500, prevalence = 0.3,
  target_hurdle_aucs = c(0.85, 0.72, 0.55),
  zero_rate_neg = c(0.55, 0.30, 0.80),
  zero_rate_pos = c(0.25, 0.10, 0.70)
)

# Check achieved zero rates vs targets
data.frame(
  Biomarker     = paste0("X", 1:3),
  Target_ZeroNeg = c(0.55, 0.30, 0.80),
  Achieved_ZeroNeg = round(sim_hur$achieved_zero_rate_neg, 3),
  Target_ZeroPos  = c(0.25, 0.10, 0.70),
  Achieved_ZeroPos = round(sim_hur$achieved_zero_rate_pos, 3)
)
```

### Advantages

- **Biologically realistic** — models the "off vs on" binary decision
  separately from the "how much when on" continuous decision
- **Recovers signal** — standard AUC near 0.5 becomes Hurdle-AUC > 0.85
  when zeros are class-discriminative
- **Interpretable** — decomposes into zero AUC and nonzero AUC components

### Limitations

- Currently independent biomarkers (no correlation control)
- Lognormal assumption for nonzero magnitudes
- Requires explicit per-class zero rates (not derived from AUC)

---

## Choosing a Simulator

| Goal | Recommended Function |
|------|---------------------|
| General-purpose matrix simulation with named correlation structures | `simulate_auc_matrix()` |
| Strong correlation preservation + accurate AUCs | `simulate_auc_copula()` |
| Zero-inflated data (scRNA-seq, microbiome) | `simulate_hurdle_auc()` |
| Checking a simulation spec is calibrated | `validate_simulation()` |
| Independent control of AUC and a fully custom correlation matrix | `generate_data_probit()` |
| Fast, simple binormal simulation | `generate_data_analytical()` |
| Exact empirical AUC for a single biomarker | `generate_auc_vector()` |
| Exact AUC + approximate Pearson correlation | `generate_auc_cor_vector()` |
| Monte Carlo sampling distribution of (AUC, r) | `simulate_auc_correlation()` |

## References

- Hanley JA, McNeil BJ (1982). The meaning and use of the area under a
  receiver operating characteristic (ROC) curve. *Radiology*, 143(1), 29–36.
- DeLong ER, DeLong DM, Clarke-Pearson DL (1988). Comparing the areas
  under two or more correlated ROC curves. *Biometrics*, 44(3), 837–845.
- Obuchowski NA (1997). Nonparametric analysis of clustered ROC curve
  data. *Biometrics*, 53(2), 567–578.
- Sun X, Xu W (2014). Fast implementation of DeLong's algorithm for
  comparing the areas under correlated ROC curves. *IEEE Signal Processing
  Letters*, 21, 1389–1393.
