---
title: "Simulating Biomarker Data in aucmat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: lightbox.html
vignette: >
  %\VignetteIndexEntry{Simulating Biomarker Data in aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r setup}
library(aucmat)
```

## Overview

`aucmat` provides three simulation engines for generating biomarker data
with controlled properties.  This vignette explains the mathematical
foundations, strengths, and limitations of each.

| Function | Approach | AUC | Correlation | Distribution |
|----------|----------|-----|-------------|-------------|
| `generate_data_probit()` | Latent multivariate normal | Target AUC via calibrated $\rho$ | Directly specified in $\Sigma$ | Multivariate normal |
| `generate_data_analytical()` | Sequential binormal decomposition | Target AUC via mean separation | Approximately achieved | Normal per class |
| `generate_auc_vector()` | Rank construction | Exact empirical AUC | Not controlled (post-hoc via Box-Cox) | Normal quantile scores |

---

## 1. Latent Probit Model (`generate_data_probit`)

### Mathematical Foundation

The binary outcome $Y$ is modelled as arising from a latent continuous
variable $Y^*$ with a threshold:

$$Y^* \sim \mathcal{N}(0, 1), \qquad
Y = \mathbf{1}\{Y^* > \tau\}, \qquad
\tau = \Phi^{-1}(1 - \pi)$$

where $\pi$ is the prevalence.

Biomarkers $(X_1, \ldots, X_p, Y^*)$ are jointly multivariate normal:

$$\begin{pmatrix} X_1 \\ \vdots \\ X_p \\ Y^* \end{pmatrix} \sim
\mathcal{N}_{p+1}\left(
\mathbf{0},
\begin{pmatrix}
\mathbf{R} & \boldsymbol{\rho} \\
\boldsymbol{\rho}^\top & 1
\end{pmatrix}
\right)$$

where $\mathbf{R}$ is the $p \times p$ between-biomarker correlation
matrix and $\rho_k = \text{Cor}(X_k, Y^*)$ governs the biomarker-outcome
relationship.

### AUC–$\rho$ Relationship

For a given prevalence $\pi$, the resulting AUC is a strictly increasing
function of the latent correlation $\rho_k$:

$$\text{AUC}(\rho_k, \pi) = f(\rho_k, \pi)$$

The function $f$ involves the bivariate normal CDF — not available in
closed form.  `generate_data_probit()` calibrates $\rho_k$ numerically
via bisection over a large calibration sample ($n = 100,000$ by default).

### Advantages

- **Both AUC and correlation are free parameters** — decoupled unlike
  the equal-variance binormal model
- **Positive definiteness** is checked and repaired automatically
- **Natural framework** — the full $(p+1) \times (p+1)$ matrix
  captures all relationships in one coherent structure

### Example

```{r probit_example}
set.seed(42)
sim_p <- generate_data_probit(
  n = 500,
  target_aucs = c(0.90, 0.80, 0.70, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.4, 0.2, 0.1,
    0.4, 1.0, 0.3, 0.2,
    0.2, 0.3, 1.0, 0.3,
    0.1, 0.2, 0.3, 1.0
  ), 4, 4),
  prevalence = 0.3
)

# Achieved AUCs closely match targets
data.frame(
  Target   = sim_p$target_aucs,
  Achieved = round(sim_p$achieved_aucs, 4)
)
```

```{r probit_latent}
# The calibrated latent correlations
sim_p$latent_rhos
```

### Limitations

- Requires `mvtnorm` for multivariate normal generation
- Calibration is computationally expensive (large $n_{\text{cal}}$)
- Not every $(p+1) \times (p+1)$ correlation matrix is positive definite —
  incompatible (AUC, correlation) specifications trigger nearest-PD repair

---

## 2. Sequential Binormal Model (`generate_data_analytical`)

### Mathematical Foundation

Under the **equal-variance binormal model**:

$$X_k \mid Y = 0 \sim \mathcal{N}(\mu_{k0}, \sigma^2), \qquad
X_k \mid Y = 1 \sim \mathcal{N}(\mu_{k1}, \sigma^2)$$

The AUC is determined by the standardized mean separation:

$$\text{AUC}_k = \Phi\!\left(\frac{\mu_{k1} - \mu_{k0}}{\sigma\sqrt{2}}\right)$$

### Sequential Construction

Biomarkers are generated from **highest to lowest AUC**.
For each biomarker $X_k$:

1. **Correlated component** $X_k^{\text{(corr)}}$ is constructed from
   the $k-1$ previously generated biomarkers to achieve the target
   correlations $\rho_{k,1}, \ldots, \rho_{k,k-1}$

2. **Residual component** $X_k^{\text{(res)}}$ carries the remaining
   mean separation needed to hit the target AUC, after accounting for
   the separation already present in the correlated component

3. $X_k = X_k^{\text{(corr)}} + X_k^{\text{(res)}}$, then standardized

### AUC–Correlation Link

In this model, for a given prevalence $\pi$, the Pearson correlation
between $X_k$ and $Y$ is **determined** by the AUC:

$$r_{pb} = \frac{\delta_k \cdot \sqrt{\pi(1-\pi)}}
                {\sqrt{\delta_k^2 \cdot \pi(1-\pi) + 1}}$$

where $\delta_k = \sqrt{2} \cdot \Phi^{-1}(\text{AUC}_k)$.

| AUC | $r_{pb}$ at $\pi = 0.3$ | $r_{pb}$ at $\pi = 0.5$ |
|-----|--------------------------|--------------------------|
| 0.60 | 0.12 | 0.16 |
| 0.70 | 0.23 | 0.31 |
| 0.80 | 0.36 | 0.47 |
| 0.90 | 0.51 | 0.64 |
| 0.95 | 0.60 | 0.73 |

### Advantages

- **Fast** — no numerical calibration needed, closed-form $\delta_k$
- **Simple** — interpretable two-class normal model
- **Good for method validation** — commonly used in biomarker simulation
  literature (Hanley & McNeil, 1982; Obuchowski, 1997)

### Limitations

- **AUC and correlation are linked** — cannot independently specify both
- **Equal variance assumption** may be restrictive for heteroscedastic
  biomarkers
- Sequential construction means biomarkers with lower AUC "inherit"
  residuals from higher-AUC biomarkers

### Example

```{r binormal_example}
set.seed(42)
sim_b <- generate_data_analytical(
  n = 500,
  prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.3, 0.1,
    0.3, 1.0, 0.2,
    0.1, 0.2, 1.0
  ), 3, 3)
)

data.frame(
  Biomarker = paste0("X", 1:3),
  Target_AUC = c(0.85, 0.75, 0.65),
  Achieved_AUC = round(sim_b$achieved_aucs, 4)
)
```

---

## 3. Rank-Based Construction (`generate_auc_vector`)

### Mathematical Foundation

The empirical AUC depends **only on the rank ordering** of scores.
On a finite sample with $n_1$ positives and $n_0$ negatives, the
achievable AUC values form a discrete grid:

$$\Delta_{\text{AUC}} = \frac{1}{n_1 \cdot n_0}$$

Given a target AUC $\theta$, we find the subset of $n_1$ ranks (from
$\{1, \ldots, n\}$) whose sum produces the required Mann-Whitney
U-statistic $U = \theta \cdot n_1 n_0$, then assign normal quantile
scores $\Phi^{-1}((R_i - 0.5)/n)$ to each observation.

### AUC–Correlation Decoupling

`generate_auc_cor_vector()` extends this by applying a strictly
increasing Box-Cox transformation after rank construction.
This preserves the AUC (ranks unchanged) while tuning the Pearson
correlation toward a target value.

```{r auc_cor}
y <- rbinom(200, 1, 0.3)
out <- generate_auc_cor_vector(
  y, target_auc = 0.80, target_cor = 0.45,
  shuffle_within_class = FALSE
)

c(AUC = out$achieved_auc, Correlation = out$achieved_cor)
```

The Box-Cox parameter $\lambda$ is found by grid search followed by
local optimization.  Not every (AUC, correlation) pair is achievable —
the monotone transform can only move the correlation within a bounded
range $[r_{\min}, r_{\max}]$.

---

## Choosing a Simulator

| Goal | Recommended Function |
|------|---------------------|
| Independent control of AUC and correlations | `generate_data_probit()` |
| Fast, simple binormal simulation | `generate_data_analytical()` |
| Exact empirical AUC for a single biomarker | `generate_auc_vector()` |
| Exact AUC + approximate Pearson correlation | `generate_auc_cor_vector()` |
| Monte Carlo sampling distribution of (AUC, r) | `simulate_auc_correlation()` |

## References

- Hanley JA, McNeil BJ (1982). The meaning and use of the area under a
  receiver operating characteristic (ROC) curve. *Radiology*, 143(1), 29–36.
- DeLong ER, DeLong DM, Clarke-Pearson DL (1988). Comparing the areas
  under two or more correlated ROC curves. *Biometrics*, 44(3), 837–845.
- Obuchowski NA (1997). Nonparametric analysis of clustered ROC curve
  data. *Biometrics*, 53(2), 567–578.
