---
title: "Simulating Biomarker Data in aucmat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      before_body: lightbox.html
vignette: >
  %\VignetteIndexEntry{Simulating Biomarker Data in aucmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r setup}
library(aucmat)
```

## Overview

`aucmat` provides several simulation engines for generating biomarker data
with controlled properties, plus a validation harness to confirm a given
specification is actually calibrated.  This vignette explains the
mathematical foundations, strengths, and limitations of each.

| Function | Approach | AUC | Correlation | Distribution |
|----------|----------|-----|-------------|-------------|
| `simulate_auc_matrix()` | Class-conditional multivariate normal | Target AUC via closed-form mean shift | Named structure (`exchangeable`/`ar1`/`block`/`user`) | Multivariate normal |
| `generate_data_probit()` | Latent multivariate normal | Target AUC via calibrated $\rho$ | Directly specified in $\Sigma$ | Multivariate normal |
| `generate_data_analytical()` | Sequential binormal decomposition | Target AUC via mean separation | Approximately achieved | Normal per class |
| `generate_auc_vector()` | Rank construction | Exact empirical AUC | Not controlled (post-hoc via Box-Cox) | Normal quantile scores |

---

## 1. Latent Probit Model (`generate_data_probit`)

### Mathematical Foundation

The binary outcome $Y$ is modelled as arising from a latent continuous
variable $Y^*$ with a threshold:

$$Y^* \sim \mathcal{N}(0, 1), \qquad
Y = \mathbf{1}\{Y^* > \tau\}, \qquad
\tau = \Phi^{-1}(1 - \pi)$$

where $\pi$ is the prevalence.

Biomarkers $(X_1, \ldots, X_p, Y^*)$ are jointly multivariate normal:

$$\begin{pmatrix} X_1 \\ \vdots \\ X_p \\ Y^* \end{pmatrix} \sim
\mathcal{N}_{p+1}\left(
\mathbf{0},
\begin{pmatrix}
\mathbf{R} & \boldsymbol{\rho} \\
\boldsymbol{\rho}^\top & 1
\end{pmatrix}
\right)$$

where $\mathbf{R}$ is the $p \times p$ between-biomarker correlation
matrix and $\rho_k = \text{Cor}(X_k, Y^*)$ governs the biomarker-outcome
relationship.

### AUC–$\rho$ Relationship

For a given prevalence $\pi$, the resulting AUC is a strictly increasing
function of the latent correlation $\rho_k$:

$$\text{AUC}(\rho_k, \pi) = f(\rho_k, \pi)$$

The function $f$ involves the bivariate normal CDF — not available in
closed form.  `generate_data_probit()` calibrates $\rho_k$ numerically
via bisection over a large calibration sample ($n = 100,000$ by default).

### Advantages

- **Both AUC and correlation are free parameters** — decoupled unlike
  the equal-variance binormal model
- **Positive definiteness** is checked and repaired automatically
- **Natural framework** — the full $(p+1) \times (p+1)$ matrix
  captures all relationships in one coherent structure

### Example

```{r probit_example}
set.seed(42)
sim_p <- generate_data_probit(
  n = 500,
  target_aucs = c(0.90, 0.80, 0.70, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.4, 0.2, 0.1,
    0.4, 1.0, 0.3, 0.2,
    0.2, 0.3, 1.0, 0.3,
    0.1, 0.2, 0.3, 1.0
  ), 4, 4),
  prevalence = 0.3
)

# Achieved AUCs closely match targets
data.frame(
  Target   = sim_p$target_aucs,
  Achieved = round(sim_p$achieved_aucs, 4)
)
```

```{r probit_latent}
# The calibrated latent correlations
sim_p$latent_rhos
```

### Limitations

- Requires `mvtnorm` for multivariate normal generation
- Calibration is computationally expensive (large $n_{\text{cal}}$)
- Not every $(p+1) \times (p+1)$ correlation matrix is positive definite —
  incompatible (AUC, correlation) specifications trigger nearest-PD repair

---

## 2. Sequential Binormal Model (`generate_data_analytical`)

### Mathematical Foundation

Under the **equal-variance binormal model**:

$$X_k \mid Y = 0 \sim \mathcal{N}(\mu_{k0}, \sigma^2), \qquad
X_k \mid Y = 1 \sim \mathcal{N}(\mu_{k1}, \sigma^2)$$

The AUC is determined by the standardized mean separation:

$$\text{AUC}_k = \Phi\!\left(\frac{\mu_{k1} - \mu_{k0}}{\sigma\sqrt{2}}\right)$$

### Sequential Construction

Biomarkers are generated from **highest to lowest AUC**.
For each biomarker $X_k$:

1. **Correlated component** $X_k^{\text{(corr)}}$ is constructed from
   the $k-1$ previously generated biomarkers to achieve the target
   correlations $\rho_{k,1}, \ldots, \rho_{k,k-1}$

2. **Residual component** $X_k^{\text{(res)}}$ carries the remaining
   mean separation needed to hit the target AUC, after accounting for
   the separation already present in the correlated component

3. $X_k = X_k^{\text{(corr)}} + X_k^{\text{(res)}}$, then standardized

### AUC–Correlation Link

In this model, for a given prevalence $\pi$, the Pearson correlation
between $X_k$ and $Y$ is **determined** by the AUC:

$$r_{pb} = \frac{\delta_k \cdot \sqrt{\pi(1-\pi)}}
                {\sqrt{\delta_k^2 \cdot \pi(1-\pi) + 1}}$$

where $\delta_k = \sqrt{2} \cdot \Phi^{-1}(\text{AUC}_k)$.

| AUC | $r_{pb}$ at $\pi = 0.3$ | $r_{pb}$ at $\pi = 0.5$ |
|-----|--------------------------|--------------------------|
| 0.60 | 0.12 | 0.16 |
| 0.70 | 0.23 | 0.31 |
| 0.80 | 0.36 | 0.47 |
| 0.90 | 0.51 | 0.64 |
| 0.95 | 0.60 | 0.73 |

### Advantages

- **Fast** — no numerical calibration needed, closed-form $\delta_k$
- **Simple** — interpretable two-class normal model
- **Good for method validation** — commonly used in biomarker simulation
  literature (Hanley & McNeil, 1982; Obuchowski, 1997)

### Limitations

- **AUC and correlation are linked** — cannot independently specify both
- **Equal variance assumption** may be restrictive for heteroscedastic
  biomarkers
- Sequential construction means biomarkers with lower AUC "inherit"
  residuals from higher-AUC biomarkers

### Example

```{r binormal_example}
set.seed(42)
sim_b <- generate_data_analytical(
  n = 500,
  prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65),
  corr_matrix = matrix(c(
    1.0, 0.3, 0.1,
    0.3, 1.0, 0.2,
    0.1, 0.2, 1.0
  ), 3, 3)
)

data.frame(
  Biomarker = paste0("X", 1:3),
  Target_AUC = c(0.85, 0.75, 0.65),
  Achieved_AUC = round(sim_b$achieved_aucs, 4)
)
```

---

## 3. Rank-Based Construction (`generate_auc_vector`)

### Mathematical Foundation

The empirical AUC depends **only on the rank ordering** of scores.
On a finite sample with $n_1$ positives and $n_0$ negatives, the
achievable AUC values form a discrete grid:

$$\Delta_{\text{AUC}} = \frac{1}{n_1 \cdot n_0}$$

Given a target AUC $\theta$, we find the subset of $n_1$ ranks (from
$\{1, \ldots, n\}$) whose sum produces the required Mann-Whitney
U-statistic $U = \theta \cdot n_1 n_0$, then assign normal quantile
scores $\Phi^{-1}((R_i - 0.5)/n)$ to each observation.

### AUC–Correlation Decoupling

`generate_auc_cor_vector()` extends this by applying a strictly
increasing Box-Cox transformation after rank construction.
This preserves the AUC (ranks unchanged) while tuning the Pearson
correlation toward a target value.

```{r auc_cor}
y <- rbinom(200, 1, 0.3)
out <- generate_auc_cor_vector(
  y, target_auc = 0.80, target_cor = 0.45,
  shuffle_within_class = FALSE
)

c(AUC = out$achieved_auc, Correlation = out$achieved_cor)
```

The Box-Cox parameter $\lambda$ is found by grid search followed by
local optimization.  Not every (AUC, correlation) pair is achievable —
the monotone transform can only move the correlation within a bounded
range $[r_{\min}, r_{\max}]$.

---

## 4. Class-Conditional Matrix Simulator (`simulate_auc_matrix`)

### Mathematical Foundation

Rather than deriving the outcome from a thresholded latent variable,
`simulate_auc_matrix()` fixes the outcome $Y$ first — supplied directly, or
generated with exact or Bernoulli class counts — and draws biomarkers
**conditional on class**:

$$X \mid Y = 0 \sim \mathcal{N}_p(\mu_0, \mathbf{R}), \qquad
X \mid Y = 1 \sim \mathcal{N}_p(\mu_1, \mathbf{R})$$

Both classes share the same correlation matrix $\mathbf{R}$; only the mean
vectors differ. The per-biomarker mean shift $\delta_k = \mu_{1k} - \mu_{0k}$
uses the same closed-form binormal calibration as
`generate_data_analytical()`:

$$\delta_k = \sqrt{2} \cdot \Phi^{-1}(\text{AUC}_k)$$

Because $Y$ is fixed *before* any biomarker is drawn, a supplied `y` keeps
its exact values and row order, and a generated `y` gets exact class counts
under the default `outcome_mode = "exact"` (or independent Bernoulli draws
under `outcome_mode = "bernoulli"`) — neither is achievable when the outcome
is instead re-derived by thresholding a fresh latent draw.

### Correlation Structures

Instead of hand-building a full $p \times p$ matrix, `structure` names one of:

| `structure` | Parameters | Off-diagonal entries |
|---|---|---|
| `"user"` | `correlation` = full matrix | as supplied |
| `"exchangeable"` | `correlation` = single value | constant $\rho$ everywhere |
| `"ar1"` | `correlation` = single value | $\rho^{|i-j|}$ |
| `"block"` | `block_sizes`, `rho_within`, `rho_between` | block-diagonal |

```{r matrix_ar1}
set.seed(1)
sim_ar1 <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.85, 0.8, 0.75),
  correlation = 0.6, structure = "ar1"
)
round(sim_ar1$achieved_correlation, 2)
```

```{r matrix_block}
sim_block <- simulate_auc_matrix(
  n = 500, prevalence = 0.3,
  target_aucs = c(0.9, 0.85, 0.75, 0.7),
  structure = "block", block_sizes = c(2, 2),
  rho_within = 0.6, rho_between = 0.1, seed = 1
)
round(sim_block$achieved_correlation, 2)
```

### Feasibility Diagnostics

Not every requested correlation matrix is positive definite (e.g. an
exchangeable matrix needs `correlation > -1/(p-1)`). `feasibility` controls
what happens when it isn't:

- **`"error"`** (default): stops with a structured `aucmat_infeasible_targets`
  condition.
- **`"nearest"`**: projects to the nearest positive-semidefinite matrix
  (eigenvalue clipping, rescaled back to a unit diagonal) and proceeds,
  reporting the adjustment in `$feasibility`.

```{r matrix_infeasible}
sim_adj <- simulate_auc_matrix(
  n = 200, prevalence = 0.3,
  target_aucs = c(0.8, 0.7, 0.6, 0.55),
  correlation = -0.5, structure = "exchangeable",
  feasibility = "nearest", seed = 1
)
sim_adj$feasibility
```

### A Note on Achieved Correlation

`$achieved_correlation` is computed on **within-class centred** residuals
(each class centred on its own empirical mean before correlating), matching
what `$requested_correlation` (`R`) actually parametrizes. The raw pooled
correlation of the returned columns — ignoring `truth` — is *not* a
consistent estimate of `R` whenever an AUC differs from 0.5: mixing two
subpopulations with different means systematically inflates the pooled
correlation (a mixture / Simpson's-paradox effect), independently of the
within-class dependence structure.

### Advantages

- **Fast** — closed-form mean shift, no numerical calibration
- **Named correlation structures** — exchangeable, AR(1), block, or a
  hand-supplied matrix
- **Feasibility diagnostics** — structured errors or automatic nearest-PD
  repair, with the adjustment magnitude reported
- **Outcome-preserving** — a supplied `y` is never altered; a generated `y`
  gets exact class counts by default

### Limitations

- Equal-variance binormal assumption, like `generate_data_analytical()`
- Correlation structure is limited to `user`/`exchangeable`/`ar1`/`block`
- Non-normal marginals are not yet supported

---

## 5. Validating a Simulator (`validate_simulation`)

A single simulated draw is not evidence that a simulator is calibrated.
`validate_simulation()` repeats a `simulate_auc_matrix()` call many times
with independent seeds and reports bias, RMSE, Monte Carlo standard error,
and target-interval hit rates for both AUCs and pairwise correlations.

```{r validate}
val <- validate_simulation(
  n = 300, prevalence = 0.3,
  target_aucs = c(0.85, 0.75, 0.65), correlation = 0.3, structure = "exchangeable",
  times = 100, seed = 1
)
print(val)
```

Bias should be small relative to the Monte Carlo SE (`auc_mc_se`); a bias
that stays large across replicates relative to `auc_mc_se` signals a
calibration bug rather than sampling noise.

---

## Choosing a Simulator

| Goal | Recommended Function |
|------|---------------------|
| General-purpose matrix simulation with named correlation structures | `simulate_auc_matrix()` |
| Checking that a simulation spec is actually calibrated | `validate_simulation()` |
| Independent control of AUC and a fully custom correlation matrix | `generate_data_probit()` |
| Fast, simple binormal simulation | `generate_data_analytical()` |
| Exact empirical AUC for a single biomarker | `generate_auc_vector()` |
| Exact AUC + approximate Pearson correlation | `generate_auc_cor_vector()` |
| Monte Carlo sampling distribution of (AUC, r) | `simulate_auc_correlation()` |

## References

- Hanley JA, McNeil BJ (1982). The meaning and use of the area under a
  receiver operating characteristic (ROC) curve. *Radiology*, 143(1), 29–36.
- DeLong ER, DeLong DM, Clarke-Pearson DL (1988). Comparing the areas
  under two or more correlated ROC curves. *Biometrics*, 44(3), 837–845.
- Obuchowski NA (1997). Nonparametric analysis of clustered ROC curve
  data. *Biometrics*, 53(2), 567–578.
